---
title: "CRE Infection Risk and Mortality in Liver Transplant Patients"
subtitle: "Analysis of the CRECOOLT Study"
author: "Data Analysis Report"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---


## Executive Summary

This report analyzes data from 692 liver transplant patients in the CRECOOLT study. Key findings: CRE infection increases mortality risk by **3.6-fold** - **71%** of CRE infections occur within the first month post-transplant - **Respiratory infections** have the highest mortality (~42%) - **Surgical/intra-abdominal infections** (SSI and IAI merged) account for 37% mortality - **Infection type matters more than baseline patient characteristics** for predicting outcomes

## Data Overview

```{r setup}
library(tidyverse)
library(survival)
library(survminer)
library(patchwork)

# Load data
CRECOOLT_overall <- read_csv("CRECOOLT_overall.csv", show_col_types = FALSE)

# Create baseline dataset (non-repeated instruments)
baseline_data <- CRECOOLT_overall |>
  filter(is.na(redcap_repeat_instrument))

# Basic summary
n_patients <- nrow(baseline_data)
n_variables <- ncol(CRECOOLT_overall)
```

The CRECOOLT dataset contains **`r n_patients` unique patients** with **`r n_variables` variables** tracking:

-   Demographics and comorbidities
-   Transplant characteristics
-   CRE colonization and infection data
-   Antibiotic resistance patterns
-   Clinical outcomes and mortality

## CRE Infection Epidemiology

### Overall Infection and Mortality Rates

```{r infection-rates}
# Calculate infection and mortality rates
infection_summary <- baseline_data |>
  summarise(
    total_patients = n(),
    cre_infections = sum(cre_infection == 1, na.rm = TRUE),
    infection_rate = round(sum(cre_infection == 1, na.rm = TRUE) / n() * 100, 1),
    overall_deaths = sum(death == 1, na.rm = TRUE),
    overall_mortality = round(sum(death == 1, na.rm = TRUE) / n() * 100, 1),
    deaths_90d = sum(mortality_90day == 1, na.rm = TRUE),
    mortality_90d = round(sum(mortality_90day == 1, na.rm = TRUE) / n() * 100, 1)
  )

knitr::kable(infection_summary, 
             caption = "Overall Infection and Mortality Statistics")
```

**Key Finding:** More than 1 in 4 liver transplant patients (`r infection_summary$infection_rate`%) develop CRE infections, with an overall mortality rate of `r infection_summary$overall_mortality`%.

### Mortality Impact of CRE Infection

```{r mortality-comparison}
# Compare mortality by CRE infection status
mortality_by_cre <- baseline_data |>
  filter(!is.na(cre_infection), !is.na(death)) |>
  group_by(cre_infection) |>
  summarise(
    n_patients = n(),
    deaths = sum(death),
    mortality_rate = round(sum(death) / n() * 100, 1)
  ) |>
  mutate(cre_status = ifelse(cre_infection == 1, "CRE Infection", "No CRE Infection"))

# Calculate risk ratio
risk_ratio <- round(mortality_by_cre$mortality_rate[2] / mortality_by_cre$mortality_rate[1], 2)

# Visualization
plot_data <- baseline_data |>
  filter(!is.na(cre_infection), !is.na(death)) |>
  mutate(
    cre_status = ifelse(cre_infection == 1, "CRE Infection", "No CRE Infection"),
    death_status = ifelse(death == 1, "Died", "Survived")
  )

ggplot(plot_data |> count(cre_status, death_status) |>
         group_by(cre_status) |> mutate(percentage = n / sum(n) * 100),
       aes(x = cre_status, y = percentage, fill = death_status)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")),
            position = position_stack(vjust = 0.5),
            color = "white", fontface = "bold", size = 5) +
  scale_fill_manual(values = c("Survived" = "#2E7D32", "Died" = "#C62828"),
                    name = "Outcome") +
  labs(title = "Mortality Rates by CRE Infection Status",
       subtitle = "Liver transplant patients (N=678)",
       x = NULL, y = "Percentage of Patients") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 16),
        legend.position = "top")

knitr::kable(mortality_by_cre |> select(cre_status, n_patients, deaths, mortality_rate),
             caption = "Mortality Comparison by CRE Infection Status")
```

**Critical Finding:** Patients with CRE infection have a **`r risk_ratio`x higher mortality rate** (36.6% vs 10.2%).

## Survival Analysis

### Kaplan-Meier Survival Curves

```{r survival-analysis, fig.height=7}
# Prepare survival data
surv_data <- baseline_data |>
  filter(!is.na(cre_infection), !is.na(death)) |>
  mutate(
    cre_status = factor(
      ifelse(cre_infection == 1, "CRE Infection", "No CRE Infection"),
      levels = c("No CRE Infection", "CRE Infection")
    ),
    time = ifelse(!is.na(OLT_to_death), OLT_to_death, 
                  ifelse(!is.na(LOS), LOS, NA)),
    event = death
  ) |>
  filter(!is.na(time), time > 0)

# Fit survival curves
surv_obj <- Surv(time = surv_data$time, event = surv_data$event)
surv_fit <- survfit(surv_obj ~ cre_status, data = surv_data)

# Plot with 360-day truncation
ggsurvplot(
  surv_fit,
  data = surv_data,
  pval = TRUE,
  conf.int = TRUE,
  risk.table = TRUE,
  risk.table.height = 0.25,
  xlab = "Time from Transplant (days)",
  ylab = "Survival Probability",
  title = "Survival Curves by CRE Infection Status",
  subtitle = "Kaplan-Meier estimates with 95% CI (1-year follow-up)",
  legend.labs = c("No CRE Infection", "CRE Infection"),
  palette = c("#2E7D32", "#C62828"),
  xlim = c(0, 360),
  break.x.by = 60,
  ggtheme = theme_minimal(base_size = 12)
)
```

### Survival at Key Timepoints

```{r survival-timepoints}
# Calculate survival at specific time points
time_points <- c(30, 90, 180)
surv_summary <- summary(surv_fit, times = time_points)

surv_rates <- data.frame(
  time = surv_summary$time,
  group = rep(c("No CRE Infection", "CRE Infection"), each = length(time_points)),
  survival_pct = round(surv_summary$surv * 100, 1),
  mortality_pct = round((1 - surv_summary$surv) * 100, 1)
) |>
  pivot_wider(names_from = group, values_from = c(survival_pct, mortality_pct))

knitr::kable(surv_rates, 
             caption = "Survival and Mortality Rates at Key Timepoints")

# Get specific values for text
surv_180_cre <- surv_rates$survival_pct_CRE_Infection[3]
surv_180_no_cre <- surv_rates$survival_pct_No_CRE_Infection[3]

# Visualization
plot_timepoints <- data.frame(
  time = rep(c("30 days", "90 days", "180 days"), each = 2),
  group = rep(c("No CRE Infection", "CRE Infection"), 3),
  survival = c(surv_summary$surv * 100)
) |>
  mutate(time = factor(time, levels = c("30 days", "90 days", "180 days")))

ggplot(plot_timepoints, aes(x = time, y = survival, fill = group)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = paste0(round(survival, 1), "%")),
            position = position_dodge(width = 0.8), vjust = -0.5, size = 4.5) +
  scale_fill_manual(values = c("No CRE Infection" = "#2E7D32", "CRE Infection" = "#C62828"),
                    name = NULL) +
  scale_y_continuous(limits = c(0, 105), breaks = seq(0, 100, 20)) +
  labs(title = "Survival Rates at Key Time Points After Transplant",
       x = "Time Point", y = "Survival Rate (%)") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")
```

By 6 months post-transplant, only **`r surv_180_cre`%** of CRE-infected patients survive, compared to **`r surv_180_no_cre`%** of non-infected patients.

## Timing of CRE Infections

### When Do Infections Occur?

```{r infection-timing}
# Analyze timing of CRE infection
cre_infected <- baseline_data |>
  filter(cre_infection == 1, !is.na(OLT_to_CREinfection))

timing_summary <- data.frame(
  Statistic = c("Median", "Mean", "25th percentile", "75th percentile"),
  Days = c(
    median(cre_infected$OLT_to_CREinfection),
    round(mean(cre_infected$OLT_to_CREinfection), 1),
    quantile(cre_infected$OLT_to_CREinfection, 0.25),
    quantile(cre_infected$OLT_to_CREinfection, 0.75)
  )
)

knitr::kable(timing_summary, 
             caption = "Time from Transplant to CRE Infection")

median_days <- median(cre_infected$OLT_to_CREinfection)

# Categorize timing
timing_categories <- cre_infected |>
  mutate(
    timing_category = case_when(
      OLT_to_CREinfection <= 7 ~ "≤7 days (1 week)",
      OLT_to_CREinfection <= 30 ~ "8-30 days (2-4 weeks)",
      OLT_to_CREinfection <= 90 ~ "31-90 days (1-3 months)",
      OLT_to_CREinfection <= 180 ~ "91-180 days (3-6 months)",
      TRUE ~ ">180 days (>6 months)"
    )
  ) |>
  count(timing_category) |>
  mutate(
    percent = round(n / sum(n) * 100, 1),
    timing_category = factor(timing_category, levels = c(
      "≤7 days (1 week)", "8-30 days (2-4 weeks)", 
      "31-90 days (1-3 months)", "91-180 days (3-6 months)", 
      ">180 days (>6 months)"
    ))
  )

# Visualization
ggplot(timing_categories, aes(x = timing_category, y = percent)) +
  geom_col(fill = "#C62828", width = 0.7) +
  geom_text(aes(label = paste0(percent, "%\n(n=", n, ")")),
            vjust = -0.3, size = 4.5, fontface = "bold") +
  scale_y_continuous(limits = c(0, 55), breaks = seq(0, 50, 10)) +
  labs(title = "When Do CRE Infections Occur After Liver Transplant?",
       subtitle = "71% of infections occur within the first month",
       x = "Time Period After Transplant",
       y = "Percentage of CRE Infections (%)") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 30, hjust = 1))

knitr::kable(timing_categories, 
             caption = "Distribution of CRE Infection Timing")
```

**Key Insight:** **71%** of CRE infections occur within the first month, with a median time to infection of just **`r median_days` days**.

## Infection Source Analysis

### Infection Type Descriptions

The following CRE infection types were recorded in this cohort:

| Infection Type | Description |
|---|---|
| **Bloodstream (BSI)** | CRE isolated from ≥1 blood culture, with or without an identified primary source. Includes primary and secondary bacteraemias. |
| **Respiratory** | CRE isolated from respiratory specimens (bronchoalveolar lavage, tracheal aspirate, or sputum) with compatible clinical signs of pneumonia or tracheobronchitis. |
| **Urinary (UTI)** | CRE isolated from urine culture with clinical signs of urinary tract infection. Includes both lower and upper urinary tract infections. |
| **Surgical/Intra-abdominal** | CRE isolated from intra-abdominal collections, surgical drains, bile, wound cultures, or peritoneal fluid after liver transplantation. Encompasses both surgical site infections (SSI) and intra-abdominal infections (IAI), which were combined due to overlapping clinical presentation and inconsistency in the original coding (see note below). |
| **Other** | CRE infections not fitting the above categories (e.g., central line-associated, skin/soft tissue). |

::: callout-warning
## Data Quality Note: SSI and IAI Coding

The original dataset recorded surgical site infections (SSI, `infection_source___5`) and intra-abdominal infections (IAI, `infection_source___4`) as separate categories. However, two anomalies suggest possible data entry errors:

1. **Implausible mortality discrepancy:** SSI showed 46.9% mortality vs 14.3% for IAI — an unusually large gap for two closely related infection foci in the same anatomical region after liver transplantation.
2. **Negative time-to-infection in the IAI group:** Some IAI patients had `OLT_to_CREinfection` values below zero (minimum: −265 days), indicating the infection was recorded as occurring *before* transplantation, which is implausible.

These findings are consistent with miscoding between SSI and IAI in a subset of patients. The two categories have therefore been **merged into a single "Surgical/Intra-abdominal" category** for all source-level analyses. No patient had both SSI and IAI coded simultaneously, so there is no double-counting.
:::

### Mortality by Infection Source

```{r infection-sources}
# Prepare infection source data
# SSI (infection_source___5) and IAI (infection_source___4) are merged into a
# single "Surgical/Intra-abdominal" category due to probable data entry errors
# (see Data Quality Note above).
infection_source_data <- baseline_data |>
  filter(cre_infection == 1) |>
  select(record_id, death, mortality_90day, starts_with("infection_source___")) |>
  mutate(
    bloodstream           = ifelse(infection_source___1 == 1, 1, 0),
    respiratory           = ifelse(infection_source___2 == 1, 1, 0),
    urinary               = ifelse(infection_source___3 == 1, 1, 0),
    surgical_abdominal    = ifelse(infection_source___4 == 1 |
                                   infection_source___5 == 1, 1, 0),
    other                 = ifelse(infection_source___6 == 1, 1, 0)
  ) |>
  filter(!is.na(death))

# Calculate mortality for each source
sources_list <- list(
  "Bloodstream (BSI)"          = "bloodstream",
  "Respiratory"                = "respiratory",
  "Urinary (UTI)"              = "urinary",
  "Surgical/Intra-abdominal"   = "surgical_abdominal",
  "Other"                      = "other"
)

mortality_by_source <- map_df(names(sources_list), function(source_name) {
  source_col <- sources_list[[source_name]]
  source_subset <- infection_source_data |>
    filter(get(source_col) == 1)
  
  if (nrow(source_subset) > 0) {
    data.frame(
      Source = source_name,
      n_infections = nrow(source_subset),
      deaths = sum(source_subset$death, na.rm = TRUE),
      mortality_rate = round(sum(source_subset$death, na.rm = TRUE) / 
                            nrow(source_subset) * 100, 1),
      deaths_90d = sum(source_subset$mortality_90day, na.rm = TRUE),
      mortality_90d_rate = round(sum(source_subset$mortality_90day, na.rm = TRUE) / 
                                nrow(source_subset) * 100, 1)
    )
  }
}) |>
  arrange(desc(mortality_rate))

highest_mort_source <- mortality_by_source$Source[1]
highest_mort <- mortality_by_source$mortality_rate[1]
lowest_mort_source <- mortality_by_source$Source[nrow(mortality_by_source)]
lowest_mort <- mortality_by_source$mortality_rate[nrow(mortality_by_source)]

knitr::kable(mortality_by_source, 
             caption = "Mortality Rates by Infection Source (SSI and IAI merged)")

# Visualization
plot_mort_source <- mortality_by_source |>
  select(Source, n_infections, mortality_90d_rate, mortality_rate) |>
  pivot_longer(cols = c(mortality_90d_rate, mortality_rate),
               names_to = "timepoint", values_to = "mortality") |>
  mutate(timepoint = factor(timepoint,
                            levels = c("mortality_90d_rate", "mortality_rate"),
                            labels = c("90-day mortality", "Overall mortality")))

ggplot(plot_mort_source, aes(x = reorder(Source, -mortality), y = mortality, fill = timepoint)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = paste0(mortality, "%")),
            position = position_dodge(width = 0.8), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("90-day mortality" = "#D32F2F", 
                                "Overall mortality" = "#9E9E9E"), name = NULL) +
  scale_y_continuous(limits = c(0, 55), breaks = seq(0, 50, 10)) +
  labs(title = "Mortality Rates by CRE Infection Source",
       subtitle = "Comparing early (90-day) vs overall mortality (SSI and IAI merged)",
       x = "Infection Source", y = "Mortality Rate (%)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 15),
        legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

**Key Findings:**

-   **`r highest_mort_source`** infections have the highest mortality (**`r highest_mort`%**)
-   **Respiratory infections** show a delayed mortality pattern (42.4% overall, but only 18.2% at 90 days), suggesting late deterioration
-   **`r lowest_mort_source`** infections have the lowest mortality (**`r lowest_mort`%**)

### Infection Sources: Survivors vs Non-Survivors

```{r source-comparison}
# Compare infection sources between survivors and non-survivors
# sources_list already uses the merged surgical_abdominal category
source_by_outcome <- map_df(names(sources_list), function(source_name) {
  source_col <- sources_list[[source_name]]
  
  survived_count <- sum(infection_source_data$death == 0 & 
                       infection_source_data[[source_col]] == 1, na.rm = TRUE)
  survived_total <- sum(infection_source_data$death == 0, na.rm = TRUE)
  
  died_count <- sum(infection_source_data$death == 1 & 
                    infection_source_data[[source_col]] == 1, na.rm = TRUE)
  died_total <- sum(infection_source_data$death == 1, na.rm = TRUE)
  
  data.frame(
    Source = source_name,
    Survived_pct = round(survived_count / survived_total * 100, 1),
    Died_pct = round(died_count / died_total * 100, 1)
  )
}) |>
  mutate(Difference = Died_pct - Survived_pct) |>
  arrange(desc(Difference))

# Diverging bar chart
ggplot(source_by_outcome, 
       aes(x = Difference, y = reorder(Source, Difference), 
           fill = ifelse(Difference > 0, "Higher in deaths", "Higher in survivors"))) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%+.1f%%", Difference)),
            hjust = ifelse(source_by_outcome$Difference > 0, -0.2, 1.2), size = 4) +
  scale_fill_manual(values = c("Higher in deaths" = "#C62828", 
                                "Higher in survivors" = "#2E7D32"), name = NULL) +
  labs(title = "Infection Source Differences: Deaths vs Survivors",
       subtitle = "Positive values indicate sources more common in patients who died (SSI and IAI merged)",
       x = "Difference in Percentage Points", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 15),
        legend.position = "top")
```

**Critical Finding:** **Respiratory infections** are the most over-represented source in patients who died. **Surgical/Intra-abdominal infections**, when SSI and IAI are properly combined, show a more moderate pattern consistent with the overall merged mortality rate of 37%.

## Patient Characteristics and Outcomes

### Baseline Characteristics by Infection Source

```{r patient-characteristics}
# Get single-source infections for cleaner comparison
# Using merged surgical_abdominal category (SSI + IAI combined)
single_source_chars <- infection_source_data |>
  filter((bloodstream + respiratory + urinary + surgical_abdominal + other) == 1) |>
  mutate(
    primary_source = case_when(
      bloodstream == 1         ~ "Bloodstream (BSI)",
      respiratory == 1         ~ "Respiratory",
      urinary == 1             ~ "Urinary (UTI)",
      surgical_abdominal == 1  ~ "Surgical/Intra-abdominal",
      TRUE ~ NA_character_
    )
  ) |>
  left_join(
    baseline_data |> select(record_id, age, gender, meld_score, cci, OLT_to_CREinfection),
    by = "record_id"
  ) |>
  filter(!is.na(primary_source), primary_source != "Other")

# MELD score comparison
meld_comparison <- single_source_chars |>
  filter(!is.na(meld_score)) |>
  group_by(primary_source) |>
  summarise(n = n(), mean_meld = round(mean(meld_score), 1), .groups = "drop") |>
  arrange(desc(mean_meld))

# Mortality rates
mortality_single <- single_source_chars |>
  filter(!is.na(death)) |>
  group_by(primary_source) |>
  summarise(n = n(), 
            deaths = sum(death),
            mortality_rate = round(sum(death) / n() * 100, 1),
            .groups = "drop")

# Timing
timing_comparison <- single_source_chars |>
  filter(!is.na(OLT_to_CREinfection)) |>
  group_by(primary_source) |>
  summarise(n = n(),
            median_days = round(median(OLT_to_CREinfection), 1),
            .groups = "drop")

# Combine for visualization
char_summary <- meld_comparison |>
  left_join(timing_comparison |> select(primary_source, median_days), by = "primary_source") |>
  left_join(mortality_single |> select(primary_source, mortality_rate), by = "primary_source")

# Multi-panel plot
p1 <- ggplot(char_summary, aes(x = reorder(primary_source, -mean_meld), y = mean_meld)) +
  geom_col(aes(fill = mortality_rate), width = 0.6) +
  geom_text(aes(label = round(mean_meld, 1)), vjust = -0.5, size = 3.5) +
  scale_fill_gradient(low = "#FFF9C4", high = "#C62828", name = "Mortality %", limits = c(0, 60)) +
  labs(title = "MELD Score", x = NULL, y = "Mean MELD") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")

p2 <- ggplot(char_summary, aes(x = reorder(primary_source, -mean_meld), y = median_days)) +
  geom_col(aes(fill = mortality_rate), width = 0.6) +
  geom_text(aes(label = round(median_days, 0)), vjust = -0.5, size = 3.5) +
  scale_fill_gradient(low = "#FFF9C4", high = "#C62828", name = "Mortality %", limits = c(0, 60)) +
  labs(title = "Time to Infection", x = NULL, y = "Median Days") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")

p3 <- ggplot(char_summary, aes(x = reorder(primary_source, -mean_meld), y = mortality_rate)) +
  geom_col(aes(fill = mortality_rate), width = 0.6) +
  geom_text(aes(label = paste0(mortality_rate, "%")), vjust = -0.5, size = 3.5) +
  scale_fill_gradient(low = "#FFF9C4", high = "#C62828", name = "Mortality %", limits = c(0, 60)) +
  labs(title = "Mortality Rate", x = NULL, y = "Mortality (%)") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

(p1 | p2 | p3) +
  plot_annotation(
    title = "Patient Characteristics by CRE Infection Source",
    subtitle = "Ordered by MELD score (disease severity) - Color intensity indicates mortality rate"
  )
```

### The MELD Paradox

```{r meld-mortality-plot, fig.height=6}
# Scatterplot showing MELD vs Mortality
ggplot(char_summary, aes(x = mean_meld, y = mortality_rate)) +
  geom_point(aes(size = n, color = primary_source), alpha = 0.8) +
  geom_text(aes(label = primary_source), vjust = -1.2, size = 3.5, fontface = "bold") +
  geom_smooth(method = "lm", se = TRUE, color = "gray40", linetype = "dashed", linewidth = 0.5) +
  scale_size_continuous(name = "N patients", range = c(5, 15)) +
  scale_color_manual(
    values = c("Surgical/Intra-abdominal" = "#D32F2F", "Respiratory" = "#F57C00",
               "Bloodstream (BSI)" = "#7B1FA2", "Urinary (UTI)" = "#388E3C"),
    name = "Infection Source"
  ) +
  labs(title = "Relationship Between Disease Severity and Mortality",
       subtitle = "MELD score at transplant vs mortality rate by infection source (SSI and IAI merged)",
       x = "Mean MELD Score (Disease Severity)", y = "Mortality Rate (%)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14), legend.position = "right")
```

**The MELD Paradox:** Patients with similar disease severity (MELD scores) have vastly different outcomes depending on infection source. With SSI and IAI merged into a single Surgical/Intra-abdominal category, the previously anomalous gap between these two groups disappears, yielding a more coherent picture:

-   **Surgical/Intra-abdominal** (n=70): ~37% mortality — a clinically plausible rate for deep abdominal CRE infections
-   **Respiratory** infections have the highest mortality despite a lower average MELD score
-   Pre-transplant disease severity (MELD) is a poor predictor of CRE infection mortality across all source types

## Key Conclusions

### 1. CRE Infection is Highly Lethal

-   CRE infection increases mortality risk by **3.6-fold**
-   Only **22.6% of CRE-infected patients survive to 6 months**
-   The impact is immediate and severe, with most deaths in the first 90 days

### 2. Timing is Critical

-   **71% of infections occur within the first month** post-transplant
-   Median time to infection: **16 days**
-   This identifies a critical surveillance and prevention window

### 3. Infection Type Matters More Than Patient Characteristics

-   **Respiratory infections** carry the highest mortality (~42%), even in patients with lower pre-transplant disease severity
-   **Surgical/Intra-abdominal infections** (SSI and IAI merged; n=70) have ~37% overall mortality — a clinically coherent rate for deep abdominal CRE infections after liver transplantation
-   **Urinary and bloodstream infections**, while still serious, have comparatively lower mortality (~36% and ~30%, respectively)
-   Pre-transplant MELD score does **NOT** predict CRE infection mortality
-   **Data quality note:** The original separate coding of SSI and IAI showed an implausible 33-percentage-point mortality difference and included negative time-to-infection values in the IAI group. These two categories have been merged for all analyses; reassessment with adjudicated source data is recommended.

### 4. Clinical Implications

**Prevention Priorities:**

1.  Focus on preventing **surgical and intra-abdominal infections** in the perioperative period through rigorous wound care, drain management, and biliary anastomosis surveillance
2.  Implement aggressive respiratory care protocols (ventilator bundles, early extubation strategies)
3.  Prompt source control for all abdominal collections

**Surveillance Strategy:**

-   Intensify monitoring in the **first 30 days** post-transplant
-   Consider infection source-specific protocols
-   Early identification enables prompt intervention

**Treatment Considerations:**

-   Infection source should guide treatment intensity
-   Respiratory CRE infections warrant special attention regardless of pre-transplant patient severity
-   Source data for SSI vs IAI should be adjudicated in future data collection to enable source-specific analyses

## Diagnostics and Treatment by Study Period

This section compares microbiological diagnostic practices and anti-CRE antibiotic therapy between the retrospective and prospective cohorts. Screening and prophylaxis comparisons use all colonised or infected patients; treatment comparisons are restricted to CRE-infected patients only.

> For the full inter-period comparison including visualisations, see the dedicated [**Diagnostics & Treatment page**](diagnostics_treatment.html).

```{r period-setup}
# ── Cohort definitions ────────────────────────────────────────────────────────
colonised_data <- baseline_data |>
  filter(cre_colon == 1 | cre_infection == 1) |>
  mutate(period = factor(ifelse(retro_or_pros == 1, "Retrospective", "Prospective"),
                         levels = c("Retrospective", "Prospective")))

infected_data <- baseline_data |>
  filter(cre_infection == 1) |>
  mutate(period = factor(ifelse(retro_or_pros == 1, "Retrospective", "Prospective"),
                         levels = c("Retrospective", "Prospective")))

# Antibiotic exposure records, joined to period and infection status
period_lkp <- baseline_data |>
  select(record_id, .period = retro_or_pros, .cre_inf = cre_infection)

cre_abx <- CRECOOLT_overall |>
  filter(redcap_repeat_instrument == "exposure_to_cre_antibiotics") |>
  select(record_id, redcap_repeat_instance, drug, combination_treat,
         indication, other_drug) |>
  left_join(period_lkp, by = "record_id") |>
  filter(!is.na(.period)) |>
  mutate(period = factor(ifelse(.period == 1, "Retrospective", "Prospective"),
                         levels = c("Retrospective", "Prospective")))

cre_abx_inf <- cre_abx |> filter(.cre_inf == 1)

drug_labels <- c("1"="Ertapenem","2"="Meropenem","3"="Amikacin",
                 "4"="Gentamicin","5"="TMP/SMX","6"="Tigecycline",
                 "7"="Colistin","8"="Ceftazidime-avibactam",
                 "9"="Meropenem/vaborbactam","10"="Imipenem/relebactam",
                 "11"="Cefiderocol","12"="Fosfomycin","13"="Other")

chi_p <- function(tbl) tryCatch(round(chisq.test(tbl)$p.value, 3), error = function(e) NA_real_)
wrs_p <- function(x, g) tryCatch(round(wilcox.test(x ~ g)$p.value, 3), error = function(e) NA_real_)

retro_n_col <- sum(colonised_data$period == "Retrospective")
pros_n_col  <- sum(colonised_data$period == "Prospective")
retro_n_inf <- sum(infected_data$period  == "Retrospective")
pros_n_inf  <- sum(infected_data$period  == "Prospective")
```

### Carbapenem Resistance Mechanisms

```{r period-resistance}
mec_labels <- c(mec_of_carbapenem_resi___1 = "KPC",
                mec_of_carbapenem_resi___2 = "OXA-48",
                mec_of_carbapenem_resi___3 = "NDM",
                mec_of_carbapenem_resi___4 = "VIM",
                mec_of_carbapenem_resi___5 = "Other")

resist_tbl <- map_df(names(mec_labels), function(v) {
  n_r <- sum(colonised_data[colonised_data$period == "Retrospective", v] == 1, na.rm = TRUE)
  n_p <- sum(colonised_data[colonised_data$period == "Prospective",   v] == 1, na.rm = TRUE)
  data.frame(
    Mechanism     = mec_labels[v],
    `Retrospective (n=244)` = paste0(n_r, " (", round(n_r / retro_n_col * 100, 1), "%)"),
    `Prospective (n=86)`    = paste0(n_p, " (", round(n_p / pros_n_col  * 100, 1), "%)"),
    p_value = chi_p(matrix(c(n_r, retro_n_col - n_r, n_p, pros_n_col - n_p), 2)),
    check.names = FALSE
  )
})

knitr::kable(resist_tbl,
  caption = "Carbapenem resistance mechanisms in colonised/infected patients by study period")
```

**Key finding:** NDM increased significantly in the prospective cohort (2.9% vs 11.6%, p = 0.004), reflecting the shifting epidemiology of carbapenem resistance mechanisms over time.

### Microbiological Screening Intensity

```{r period-screening}
screen_vars   <- c(blood = "micro_screen_blood", resp = "micro_screen_resp",
                   urine = "micro_screen_urino",  abd  = "micro_screen_intra_abd")
screen_labels <- c(blood = "Blood cultures", resp = "Respiratory samples",
                   urine = "Urine cultures",  abd  = "Intra-abdominal samples")

screen_tbl <- map_df(names(screen_vars), function(k) {
  x <- colonised_data[[screen_vars[k]]]
  g <- colonised_data$period
  r <- x[g == "Retrospective"]; p <- x[g == "Prospective"]
  data.frame(
    `Sample type` = screen_labels[k],
    `Retrospective [median (IQR)]` = paste0(median(r, na.rm=T), " (",
                                             quantile(r, .25, na.rm=T), "–",
                                             quantile(r, .75, na.rm=T), ")"),
    `Prospective [median (IQR)]`   = paste0(median(p, na.rm=T), " (",
                                             quantile(p, .25, na.rm=T), "–",
                                             quantile(p, .75, na.rm=T), ")"),
    p_value = wrs_p(x, g),
    check.names = FALSE
  )
})

knitr::kable(screen_tbl,
  caption = "Microbiological screening intensity by study period (colonised/infected patients)")
```

**Key finding:** Blood culture and intra-abdominal sampling intensity increased significantly in the prospective period (blood: 5 vs 9 per patient, p = 0.002; intra-abdominal: 2 vs 3, p = 0.007).

### Colonisation Management Strategy

```{r period-strategy}
strat_labels <- c(cre_colonization___1 = "No strategy adopted",
                  cre_colonization___2 = "Targeted peri-operative prophylaxis",
                  cre_colonization___3 = "Pre-emptive strategy")

strat_tbl <- map_df(names(strat_labels), function(v) {
  n_r <- sum(colonised_data[colonised_data$period == "Retrospective", v] == 1, na.rm = TRUE)
  n_p <- sum(colonised_data[colonised_data$period == "Prospective",   v] == 1, na.rm = TRUE)
  data.frame(
    Strategy = strat_labels[v],
    `Retrospective (n=244)` = paste0(n_r, " (", round(n_r / retro_n_col * 100, 1), "%)"),
    `Prospective (n=86)`    = paste0(n_p, " (", round(n_p / pros_n_col  * 100, 1), "%)"),
    p_value = chi_p(matrix(c(n_r, retro_n_col - n_r, n_p, pros_n_col - n_p), 2)),
    check.names = FALSE
  )
})

knitr::kable(strat_tbl,
  caption = "Colonisation management strategy by study period (colonised/infected patients)")
```

### Anti-CRE Therapy in Infected Patients

#### Agents Used

```{r period-drugs}
drug_tbl <- cre_abx_inf |>
  filter(!is.na(drug)) |>
  mutate(drug_name = drug_labels[as.character(drug)]) |>
  count(period, drug_name) |>
  pivot_wider(names_from = period, values_from = n, values_fill = 0) |>
  mutate(
    p_value = map2_dbl(Retrospective, Prospective,
      ~ chi_p(matrix(c(.x, retro_n_inf - .x, .y, pros_n_inf - .y), 2))),
    Retrospective = paste0(Retrospective, " (", round(Retrospective / retro_n_inf * 100, 1), "%)"),
    Prospective   = paste0(Prospective,   " (", round(Prospective   / pros_n_inf  * 100, 1), "%)")
  ) |>
  arrange(drug_name) |>
  rename(Drug = drug_name,
         `Retrospective (n=136)` = Retrospective,
         `Prospective (n=54)`    = Prospective)

knitr::kable(drug_tbl,
  caption = "Anti-CRE agents used in infected patients by study period (exposure-level; patients may receive >1 agent)")
```

**Key findings:**

- **Ceftazidime-avibactam** use increased markedly (55.9% → 77.8%, p = 0.008), reflecting its emergence as the preferred backbone agent for KPC-producers
- **Colistin** use declined substantially (34.6% → 11.1%, p = 0.002), consistent with the shift away from polymyxin-based regimens
- **Fosfomycin** use also declined (12.5% → 1.9%, p = 0.047)

#### Combination Therapy and Treatment Indication

```{r period-combo-indication}
# Combination therapy — patient level
combo_pts <- cre_abx_inf |>
  filter(!is.na(combination_treat)) |>
  distinct(record_id, .keep_all = TRUE)

n_cr <- sum(combo_pts$period == "Retrospective" & combo_pts$combination_treat == 1)
n_cp <- sum(combo_pts$period == "Prospective"   & combo_pts$combination_treat == 1)
nr_c <- sum(combo_pts$period == "Retrospective")
np_c <- sum(combo_pts$period == "Prospective")

combo_tbl <- data.frame(
  Metric = "Combination therapy",
  `Retrospective (n=136)` = paste0(n_cr, " (", round(n_cr / nr_c * 100, 1), "%)"),
  `Prospective (n=54)`    = paste0(n_cp, " (", round(n_cp / np_c * 100, 1), "%)"),
  p_value = chi_p(matrix(c(n_cr, nr_c - n_cr, n_cp, np_c - n_cp), 2)),
  check.names = FALSE
)

# Indication
ind_wide <- cre_abx_inf |>
  filter(!is.na(indication)) |>
  mutate(ind_label = case_when(
    indication == 1 ~ "Empiric (symptoms onset)",
    indication == 2 ~ "CRECOOLT score-guided",
    indication == 3 ~ "Targeted (documented infection)"
  )) |>
  count(period, ind_label) |>
  group_by(period) |>
  mutate(pct = round(n / sum(n) * 100, 1),
         label = paste0(n, " (", pct, "%)")) |>
  ungroup() |>
  select(period, ind_label, label) |>
  pivot_wider(names_from = period, values_from = label, values_fill = "0 (0%)") |>
  rename(Metric = ind_label,
         `Retrospective (n=136)` = Retrospective,
         `Prospective (n=54)`    = Prospective) |>
  mutate(p_value = NA_real_)

bind_rows(combo_tbl, ind_wide) |>
  knitr::kable(caption = "Combination therapy and treatment indication by study period")
```

**Key findings:**

- **Combination therapy declined sharply** between periods (73.1% → 39.2%, p < 0.001), consistent with the transition to novel β-lactam/β-lactamase inhibitor agents that are effective as monotherapy
- **CRECOOLT score-guided therapy** appeared for the first time in the prospective cohort (5.5%), reflecting real-world adoption of the prediction model

## Technical Notes

**Data Source:** CRECOOLT study database (N=692 liver transplant patients)

**Statistical Methods:**

-   Kaplan-Meier survival analysis with log-rank test
-   Descriptive statistics and group comparisons; chi-square test for proportions, Wilcoxon rank-sum test for continuous variables
-   Risk ratio calculations

**Software:** R version 4.5.1, with tidyverse, survival, and survminer packages

**Report Generated:** `r format(Sys.Date(), "%B %d, %Y")`