---
title: "Diagnostics and Treatment by Study Period"
subtitle: "Retrospective vs Prospective Cohort — CRECOOLT Study"
author: "Data Analysis Report"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This report compares microbiological diagnostic practices, carbapenem resistance epidemiology, and anti-CRE antibiotic therapy between the **retrospective** (n = 521) and **prospective** (n = 171) cohorts of the CRECOOLT study.

::: {.callout-note}
## Scope

- **Screening, resistance mechanisms, and prophylaxis/management strategy:** all colonised or infected patients (retrospective n = 244; prospective n = 86)
- **Anti-CRE treatment comparisons:** CRE-infected patients only (retrospective n = 136; prospective n = 54)
:::

```{r setup}
library(tidyverse)
library(patchwork)

# Load data
CRECOOLT_overall <- read_csv("CRECOOLT_overall.csv", show_col_types = FALSE)
baseline_data <- CRECOOLT_overall |> filter(is.na(redcap_repeat_instrument))

# ── Cohort definitions ────────────────────────────────────────────────────────
colonised_data <- baseline_data |>
  filter(cre_colon == 1 | cre_infection == 1) |>
  mutate(period = factor(ifelse(retro_or_pros == 1, "Retrospective", "Prospective"),
                         levels = c("Retrospective", "Prospective")))

infected_data <- baseline_data |>
  filter(cre_infection == 1) |>
  mutate(period = factor(ifelse(retro_or_pros == 1, "Retrospective", "Prospective"),
                         levels = c("Retrospective", "Prospective")))

# ── Antibiotic exposure data ──────────────────────────────────────────────────
# Select only required columns before joining to avoid name collisions with
# retro_or_pros / cre_infection in the repeated-instrument rows (which are NA)
period_lkp <- baseline_data |>
  select(record_id, .period = retro_or_pros, .cre_inf = cre_infection)

cre_abx <- CRECOOLT_overall |>
  filter(redcap_repeat_instrument == "exposure_to_cre_antibiotics") |>
  select(record_id, redcap_repeat_instance, drug, combination_treat,
         indication, other_drug) |>
  left_join(period_lkp, by = "record_id") |>
  filter(!is.na(.period)) |>
  mutate(period = factor(ifelse(.period == 1, "Retrospective", "Prospective"),
                         levels = c("Retrospective", "Prospective")))

cre_abx_inf <- cre_abx |> filter(.cre_inf == 1)

# ── Constants and helpers ─────────────────────────────────────────────────────
drug_labels <- c(
  "1" = "Ertapenem",        "2"  = "Meropenem",
  "3" = "Amikacin",         "4"  = "Gentamicin",
  "5" = "TMP/SMX",          "6"  = "Tigecycline",
  "7" = "Colistin",         "8"  = "Ceftazidime-avibactam",
  "9" = "Meropenem/vaborbactam", "10" = "Imipenem/relebactam",
  "11" = "Cefiderocol",     "12" = "Fosfomycin",
  "13" = "Other"
)

chi_p <- function(tbl) tryCatch(round(chisq.test(tbl)$p.value, 3), error = function(e) NA_real_)
wrs_p <- function(x, g) tryCatch(round(wilcox.test(x ~ g)$p.value, 3), error = function(e) NA_real_)

retro_n_col <- sum(colonised_data$period == "Retrospective")
pros_n_col  <- sum(colonised_data$period == "Prospective")
retro_n_inf <- sum(infected_data$period  == "Retrospective")
pros_n_inf  <- sum(infected_data$period  == "Prospective")
```

## Carbapenem Resistance Mechanisms

Resistance mechanisms were characterised from the CRE colonising strain in all colonised/infected patients.

```{r resistance-table}
mec_labels <- c(
  mec_of_carbapenem_resi___1 = "KPC",
  mec_of_carbapenem_resi___2 = "OXA-48",
  mec_of_carbapenem_resi___3 = "NDM",
  mec_of_carbapenem_resi___4 = "VIM",
  mec_of_carbapenem_resi___5 = "Other"
)

resist_tbl <- map_df(names(mec_labels), function(v) {
  n_r <- sum(colonised_data[colonised_data$period == "Retrospective", v] == 1, na.rm = TRUE)
  n_p <- sum(colonised_data[colonised_data$period == "Prospective",   v] == 1, na.rm = TRUE)
  data.frame(
    Mechanism             = mec_labels[v],
    `Retrospective (n=244)` = paste0(n_r, " (", round(n_r / retro_n_col * 100, 1), "%)"),
    `Prospective (n=86)`    = paste0(n_p, " (", round(n_p / pros_n_col  * 100, 1), "%)"),
    p_value = chi_p(matrix(c(n_r, retro_n_col - n_r, n_p, pros_n_col - n_p), 2)),
    check.names = FALSE
  )
})

knitr::kable(resist_tbl,
  caption = "Carbapenem resistance mechanisms in colonised/infected patients by study period")
```

```{r resistance-plot, fig.cap="Carbapenem resistance mechanisms by study period"}
resist_plot_data <- map_df(names(mec_labels), function(v) {
  n_r <- sum(colonised_data[colonised_data$period == "Retrospective", v] == 1, na.rm = TRUE)
  n_p <- sum(colonised_data[colonised_data$period == "Prospective",   v] == 1, na.rm = TRUE)
  data.frame(
    Mechanism = mec_labels[v],
    Retrospective = n_r / retro_n_col * 100,
    Prospective   = n_p / pros_n_col  * 100
  )
}) |>
  pivot_longer(cols = c(Retrospective, Prospective),
               names_to = "Period", values_to = "Percent") |>
  mutate(Period = factor(Period, levels = c("Retrospective", "Prospective")))

ggplot(resist_plot_data, aes(x = Mechanism, y = Percent, fill = Period)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_text(aes(label = paste0(round(Percent, 1), "%")),
            position = position_dodge(width = 0.75), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Retrospective" = "#1565C0", "Prospective" = "#E53935"),
                    name = NULL) +
  scale_y_continuous(limits = c(0, 85), breaks = seq(0, 80, 20)) +
  labs(title = "Carbapenem Resistance Mechanisms by Study Period",
       subtitle = "Colonised/infected patients (retrospective n=244; prospective n=86)",
       x = "Resistance Mechanism", y = "Prevalence (%)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 15),
        legend.position = "top")
```

**Key finding:** NDM increased significantly in the prospective cohort (2.9% → 11.6%, p = 0.004), suggesting a change in the predominant resistance epidemiology. KPC remained the most common mechanism in both periods.

## Microbiological Screening Intensity

Numbers of microbiological samples collected per patient during the follow-up period.

```{r screening-table}
screen_vars   <- c(blood = "micro_screen_blood",   resp  = "micro_screen_resp",
                   urine = "micro_screen_urino",    abd   = "micro_screen_intra_abd")
screen_labels <- c(blood = "Blood cultures",        resp  = "Respiratory samples",
                   urine = "Urine cultures",         abd   = "Intra-abdominal samples")

screen_tbl <- map_df(names(screen_vars), function(k) {
  x <- colonised_data[[screen_vars[k]]]
  g <- colonised_data$period
  r <- x[g == "Retrospective"]; p <- x[g == "Prospective"]
  data.frame(
    `Sample type` = screen_labels[k],
    `Retrospective [median (IQR)]` = paste0(
      median(r, na.rm=T), " (", quantile(r, .25, na.rm=T), "–", quantile(r, .75, na.rm=T), ")"),
    `Prospective [median (IQR)]` = paste0(
      median(p, na.rm=T), " (", quantile(p, .25, na.rm=T), "–", quantile(p, .75, na.rm=T), ")"),
    p_value = wrs_p(x, g),
    check.names = FALSE
  )
})

knitr::kable(screen_tbl,
  caption = "Microbiological screening intensity by study period (colonised/infected patients)")
```

```{r screening-plot, fig.cap="Microbiological screening intensity by study period"}
screen_plot <- map_df(names(screen_vars), function(k) {
  df <- colonised_data[, c("period", screen_vars[k])]
  names(df)[2] <- "n_samples"
  df$sample <- screen_labels[k]
  df
}) |> filter(!is.na(n_samples))

ggplot(screen_plot, aes(x = period, y = n_samples, fill = period)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1.5, alpha = 0.8) +
  facet_wrap(~sample, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = c("Retrospective" = "#1565C0", "Prospective" = "#E53935"),
                    name = NULL) +
  labs(title = "Microbiological Screening Intensity by Study Period",
       subtitle = "Colonised/infected patients — number of samples per patient",
       x = NULL, y = "Number of samples") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

**Key finding:** Blood culture and intra-abdominal sampling intensity increased significantly in the prospective period (blood: median 5 vs 9, p = 0.002; intra-abdominal: 2 vs 3, p = 0.007), likely reflecting more structured surveillance protocols in the prospective phase.

## Colonisation Management Strategy

```{r strategy-table}
strat_labels <- c(
  cre_colonization___1 = "No strategy adopted",
  cre_colonization___2 = "Targeted peri-operative prophylaxis",
  cre_colonization___3 = "Pre-emptive strategy"
)

strat_tbl <- map_df(names(strat_labels), function(v) {
  n_r <- sum(colonised_data[colonised_data$period == "Retrospective", v] == 1, na.rm = TRUE)
  n_p <- sum(colonised_data[colonised_data$period == "Prospective",   v] == 1, na.rm = TRUE)
  data.frame(
    Strategy              = strat_labels[v],
    `Retrospective (n=244)` = paste0(n_r, " (", round(n_r / retro_n_col * 100, 1), "%)"),
    `Prospective (n=86)`    = paste0(n_p, " (", round(n_p / pros_n_col  * 100, 1), "%)"),
    p_value = chi_p(matrix(c(n_r, retro_n_col - n_r, n_p, pros_n_col - n_p), 2)),
    check.names = FALSE
  )
})

knitr::kable(strat_tbl,
  caption = "Colonisation management strategy by study period (colonised/infected patients)")
```

```{r strategy-plot, fig.cap="Colonisation management strategy by study period", fig.height=5}
strat_plot_data <- map_df(names(strat_labels), function(v) {
  n_r <- sum(colonised_data[colonised_data$period == "Retrospective", v] == 1, na.rm = TRUE)
  n_p <- sum(colonised_data[colonised_data$period == "Prospective",   v] == 1, na.rm = TRUE)
  data.frame(
    Strategy = strat_labels[v],
    Retrospective = n_r / retro_n_col * 100,
    Prospective   = n_p / pros_n_col  * 100
  )
}) |>
  pivot_longer(cols = c(Retrospective, Prospective),
               names_to = "Period", values_to = "Percent") |>
  mutate(Period = factor(Period, levels = c("Retrospective", "Prospective")))

ggplot(strat_plot_data, aes(x = reorder(Strategy, -Percent), y = Percent, fill = Period)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_text(aes(label = paste0(round(Percent, 1), "%")),
            position = position_dodge(width = 0.75), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Retrospective" = "#1565C0", "Prospective" = "#E53935"),
                    name = NULL) +
  scale_y_continuous(limits = c(0, 75), breaks = seq(0, 70, 10)) +
  labs(title = "Colonisation Management Strategy by Study Period",
       subtitle = "Colonised/infected patients",
       x = NULL, y = "Percentage of patients (%)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 15),
        legend.position = "top",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

## Anti-CRE Treatment in Infected Patients

### Agents Used

Exposure-level data: a single patient may have received more than one agent.

```{r drugs-table}
drug_tbl <- cre_abx_inf |>
  filter(!is.na(drug)) |>
  mutate(drug_name = drug_labels[as.character(drug)]) |>
  count(period, drug_name) |>
  pivot_wider(names_from = period, values_from = n, values_fill = 0) |>
  mutate(
    p_value = map2_dbl(Retrospective, Prospective,
      ~ chi_p(matrix(c(.x, retro_n_inf - .x, .y, pros_n_inf - .y), 2))),
    Retrospective = paste0(Retrospective, " (", round(Retrospective / retro_n_inf * 100, 1), "%)"),
    Prospective   = paste0(Prospective,   " (", round(Prospective   / pros_n_inf  * 100, 1), "%)")
  ) |>
  arrange(drug_name) |>
  rename(Drug = drug_name,
         `Retrospective (n=136)` = Retrospective,
         `Prospective (n=54)`    = Prospective)

knitr::kable(drug_tbl,
  caption = "Anti-CRE agents used in infected patients by study period (patients may receive >1 agent)")
```

```{r drugs-plot, fig.height=6, fig.cap="Anti-CRE drug use by study period (infected patients)"}
drug_plot_data <- cre_abx_inf |>
  filter(!is.na(drug)) |>
  mutate(drug_name = drug_labels[as.character(drug)]) |>
  count(period, drug_name) |>
  mutate(pct = ifelse(period == "Retrospective",
                      n / retro_n_inf * 100,
                      n / pros_n_inf  * 100)) |>
  filter(drug_name != "Other")

# Highlight significantly changed drugs
sig_drugs <- c("Ceftazidime-avibactam", "Colistin", "Fosfomycin")

ggplot(drug_plot_data,
       aes(x = reorder(drug_name, pct), y = pct, fill = period)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_text(aes(label = paste0(round(pct, 0), "%")),
            position = position_dodge(width = 0.75), hjust = -0.15, size = 3.2) +
  scale_fill_manual(values = c("Retrospective" = "#1565C0", "Prospective" = "#E53935"),
                    name = NULL) +
  scale_y_continuous(limits = c(0, 95), breaks = seq(0, 90, 20)) +
  coord_flip() +
  labs(title = "Anti-CRE Drug Use by Study Period",
       subtitle = "CRE-infected patients (retrospective n=136; prospective n=54)",
       x = NULL, y = "% of infected patients receiving agent") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 15),
        legend.position = "top")
```

**Key findings:**

- **Ceftazidime-avibactam** use increased markedly (55.9% → 77.8%, p = 0.008), reflecting its emergence as the preferred backbone agent for KPC-producers
- **Colistin** use declined substantially (34.6% → 11.1%, p = 0.002), consistent with the shift away from polymyxin-based regimens given their toxicity profile
- **Fosfomycin** use also declined (12.5% → 1.9%, p = 0.047), likely replaced by newer β-lactam/β-lactamase inhibitor combinations
- **Meropenem** remained in common use in both periods (~68% vs 63%), typically as part of combination regimens or for susceptible isolates

### Combination Therapy and Treatment Indication

```{r combo-indication}
# Combination therapy — patient level (first record per patient)
combo_pts <- cre_abx_inf |>
  filter(!is.na(combination_treat)) |>
  distinct(record_id, .keep_all = TRUE)

n_cr <- sum(combo_pts$period == "Retrospective" & combo_pts$combination_treat == 1)
n_cp <- sum(combo_pts$period == "Prospective"   & combo_pts$combination_treat == 1)
nr_c <- sum(combo_pts$period == "Retrospective")
np_c <- sum(combo_pts$period == "Prospective")

combo_row <- data.frame(
  Metric                  = "Combination therapy",
  `Retrospective (n=136)` = paste0(n_cr, " (", round(n_cr / nr_c * 100, 1), "%)"),
  `Prospective (n=54)`    = paste0(n_cp, " (", round(n_cp / np_c * 100, 1), "%)"),
  p_value = chi_p(matrix(c(n_cr, nr_c - n_cr, n_cp, np_c - n_cp), 2)),
  check.names = FALSE
)

# Indication
ind_wide <- cre_abx_inf |>
  filter(!is.na(indication)) |>
  mutate(ind_label = case_when(
    indication == 1 ~ "Empiric (symptoms onset)",
    indication == 2 ~ "CRECOOLT score-guided",
    indication == 3 ~ "Targeted (documented infection)"
  )) |>
  count(period, ind_label) |>
  group_by(period) |>
  mutate(pct   = round(n / sum(n) * 100, 1),
         label = paste0(n, " (", pct, "%)")) |>
  ungroup() |>
  select(period, ind_label, label) |>
  pivot_wider(names_from = period, values_from = label, values_fill = "0 (0%)") |>
  rename(Metric                  = ind_label,
         `Retrospective (n=136)` = Retrospective,
         `Prospective (n=54)`    = Prospective) |>
  mutate(p_value = NA_real_)

bind_rows(combo_row, ind_wide) |>
  knitr::kable(caption = "Combination therapy and treatment indication by study period")
```

```{r combo-plot, fig.height=5, fig.cap="Combination therapy and indication by study period"}
# Combination therapy bar
combo_df <- data.frame(
  Period    = c("Retrospective", "Prospective"),
  Combo_pct = c(round(n_cr / nr_c * 100, 1), round(n_cp / np_c * 100, 1))
) |> mutate(Period = factor(Period, levels = c("Retrospective", "Prospective")))

p_combo <- ggplot(combo_df, aes(x = Period, y = Combo_pct, fill = Period)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0(Combo_pct, "%")), vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("Retrospective" = "#1565C0", "Prospective" = "#E53935")) +
  scale_y_continuous(limits = c(0, 90), breaks = seq(0, 80, 20)) +
  labs(title = "Combination Therapy", x = NULL, y = "% patients") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", plot.title = element_text(face = "bold"))

# Indication stacked bar
ind_plot_data <- cre_abx_inf |>
  filter(!is.na(indication)) |>
  mutate(ind_label = case_when(
    indication == 1 ~ "Empiric",
    indication == 2 ~ "Score-guided",
    indication == 3 ~ "Targeted"
  )) |>
  count(period, ind_label) |>
  group_by(period) |>
  mutate(pct = n / sum(n) * 100)

p_ind <- ggplot(ind_plot_data, aes(x = period, y = pct, fill = ind_label)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0(round(pct, 0), "%")),
            position = position_stack(vjust = 0.5), color = "white",
            fontface = "bold", size = 4) +
  scale_fill_manual(
    values = c("Empiric" = "#EF6C00", "Score-guided" = "#2E7D32", "Targeted" = "#4527A0"),
    name = "Indication") +
  scale_y_continuous(breaks = seq(0, 100, 25)) +
  labs(title = "Treatment Indication", x = NULL, y = "% antibiotic exposures") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "right")

p_combo | p_ind +
  plot_annotation(
    title    = "Treatment Patterns by Study Period",
    subtitle = "CRE-infected patients"
  )
```

**Key findings:**

- **Combination therapy declined sharply** between periods (73.1% → 39.2%, p < 0.001), consistent with the transition to novel β-lactam/β-lactamase inhibitor agents that are effective as monotherapy
- **CRECOOLT score-guided therapy** appeared for the first time in the prospective cohort (5.5%), reflecting real-world adoption of the prediction model in clinical decision-making

## Summary

```{r summary-table}
summary_df <- data.frame(
  Finding = c(
    "NDM resistance mechanism",
    "Blood culture intensity (median)",
    "Intra-abdominal sampling (median)",
    "Ceftazidime-avibactam use",
    "Colistin use",
    "Combination therapy"
  ),
  Retrospective = c("2.9%", "5 (IQR 2–11)", "2 (IQR 1–4)", "55.9%", "34.6%", "73.1%"),
  Prospective   = c("11.6%", "9 (IQR 4–16)", "3 (IQR 1.75–6)", "77.8%", "11.1%", "39.2%"),
  p_value       = c("0.004", "0.002", "0.007", "0.008", "0.002", "<0.001"),
  Direction     = c("↑", "↑", "↑", "↑", "↓", "↓")
)

knitr::kable(summary_df,
  caption = "Summary of statistically significant changes between study periods")
```

## Survival Analysis by CRE Therapy and Resistance Mechanism

Kaplan-Meier survival curves from time of CRE infection to death, censored at hospital discharge or 90 days. Patients are classified by:

- **Therapy group:** any exposure to ceftazidime-avibactam, meropenem-vaborbactam, or cefiderocol ("Novel BL/BLI") vs other agents
- **Carbapenemase mechanism:** KPC vs Non-KPC (OXA-48, NDM, VIM, or other)

```{r km-setup}
library(survival)
library(survminer)

# ── JAMA-style theme ──────────────────────────────────────────────────────────
jama_theme <- theme_classic(base_size = 11) +
  theme(
    plot.title         = element_text(face = "bold", size = 12, hjust = 0),
    plot.subtitle      = element_text(size = 10, hjust = 0, color = "grey40"),
    axis.title         = element_text(size = 10),
    axis.text          = element_text(size = 9),
    legend.title       = element_blank(),
    legend.text        = element_text(size = 9),
    legend.position    = "top",
    legend.key.width   = unit(1.2, "cm"),
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.3),
    strip.background   = element_blank(),
    strip.text         = element_text(face = "bold")
  )

# ── Therapy classification per patient ───────────────────────────────────────
# Novel BL/BLI = any exposure to ceftazidime-avibactam (8),
#                meropenem-vaborbactam (9), or cefiderocol (11)
therapy_per_patient <- cre_abx_inf |>
  group_by(record_id) |>
  summarise(
    new_therapy  = any(drug %in% c(8, 9, 11), na.rm = TRUE),
    combination  = any(combination_treat == 1, na.rm = TRUE),
    .groups = "drop"
  )

# ── Survival dataset ──────────────────────────────────────────────────────────
surv_infected <- infected_data |>
  left_join(therapy_per_patient, by = "record_id") |>
  mutate(
    surv_time = case_when(
      !is.na(death_date) & !is.na(cre_infection_date) ~
        as.numeric(death_date - cre_infection_date),
      !is.na(discharge_date) & !is.na(cre_infection_date) ~
        as.numeric(discharge_date - cre_infection_date),
      TRUE ~ NA_real_
    ),
    surv_event = ifelse(death == 1, 1, 0),
    surv_time  = ifelse(surv_time < 0, NA_real_, surv_time),

    therapy_group = factor(
      ifelse(new_therapy, "Novel BL/BLI", "Other therapies"),
      levels = c("Novel BL/BLI", "Other therapies")
    ),
    combo_group = factor(
      case_when(
        new_therapy    ~ "Novel BL/BLI",
        combination    ~ "Combination (other agents)",
        TRUE           ~ "Monotherapy (other agents)"
      ),
      levels = c("Novel BL/BLI", "Combination (other agents)",
                 "Monotherapy (other agents)")
    ),
    carbapenemase = case_when(
      cre_isolate___1 == 1 ~ "KPC",
      cre_isolate___2 == 1 ~ "OXA-48",
      cre_isolate___3 == 1 ~ "VIM",
      cre_isolate___4 == 1 ~ "NDM",
      cre_isolate___5 == 1 ~ "Other/Unknown",
      TRUE                 ~ "Not typed"
    ),
    carba_group = factor(
      ifelse(carbapenemase == "KPC",
             "KPC",
             "Non-KPC (OXA-48/NDM/VIM/Other)"),
      levels = c("KPC", "Non-KPC (OXA-48/NDM/VIM/Other)")
    )
  ) |>
  filter(!is.na(surv_time), surv_time >= 0, !is.na(surv_event))
```

### Novel BL/BLI vs Other Therapies {#fig-km1}

```{r km-plot1, fig.height=5.5, fig.cap="Kaplan-Meier survival curves: novel BL/BLI vs other therapies"}
fit1 <- survfit(Surv(surv_time, surv_event) ~ therapy_group,
                data = surv_infected)

ggsurvplot(
  fit1,
  data              = surv_infected,
  pval              = TRUE,
  pval.size         = 3.5,
  conf.int          = TRUE,
  conf.int.alpha    = 0.12,
  risk.table        = TRUE,
  risk.table.height = 0.22,
  risk.table.title  = "No. at risk",
  risk.table.fontsize = 3,
  xlab              = "Days from CRE Infection",
  ylab              = "Overall Survival",
  title             = "A  Novel BL/BLI vs Other Therapies",
  subtitle          = "CRE-infected patients (n=165); censored at discharge or 90 days",
  legend.labs       = c("Novel BL/BLI (CAZ-AVI / MER-VAB / CEF)",
                        "Other therapies"),
  palette           = c("#374E55", "#DF8F44"),
  xlim              = c(0, 90),
  break.x.by        = 15,
  ylim              = c(0, 1),
  surv.scale        = "percent",
  ggtheme           = jama_theme,
  tables.theme      = theme_cleantable()
)
```

### Novel BL/BLI vs Combination vs Monotherapy {#fig-km2}

```{r km-plot2, fig.height=6, fig.cap="Kaplan-Meier survival curves: novel BL/BLI vs combination vs monotherapy with other agents"}
fit2 <- survfit(Surv(surv_time, surv_event) ~ combo_group,
                data = surv_infected)

ggsurvplot(
  fit2,
  data              = surv_infected,
  pval              = TRUE,
  pval.size         = 3.5,
  conf.int          = TRUE,
  conf.int.alpha    = 0.12,
  risk.table        = TRUE,
  risk.table.height = 0.25,
  risk.table.title  = "No. at risk",
  risk.table.fontsize = 3,
  xlab              = "Days from CRE Infection",
  ylab              = "Overall Survival",
  title             = "B  Novel BL/BLI vs Combination vs Monotherapy",
  subtitle          = "Combination and monotherapy groups exclude novel BL/BLI agents",
  legend.labs       = c("Novel BL/BLI",
                        "Combination (other agents)",
                        "Monotherapy (other agents)"),
  palette           = c("#374E55", "#DF8F44", "#B24745"),
  xlim              = c(0, 90),
  break.x.by        = 15,
  ylim              = c(0, 1),
  surv.scale        = "percent",
  ggtheme           = jama_theme,
  tables.theme      = theme_cleantable()
)
```

### Survival by Carbapenemase Mechanism {#fig-km3}

```{r km-plot3, fig.height=5.5, fig.cap="Kaplan-Meier survival curves by carbapenemase mechanism (KPC vs Non-KPC)"}
surv_typed <- surv_infected |> filter(carbapenemase != "Not typed")
fit3 <- survfit(Surv(surv_time, surv_event) ~ carba_group,
                data = surv_typed)

ggsurvplot(
  fit3,
  data              = surv_typed,
  pval              = TRUE,
  pval.size         = 3.5,
  conf.int          = TRUE,
  conf.int.alpha    = 0.12,
  risk.table        = TRUE,
  risk.table.height = 0.22,
  risk.table.title  = "No. at risk",
  risk.table.fontsize = 3,
  xlab              = "Days from CRE Infection",
  ylab              = "Overall Survival",
  title             = "C  Survival by Carbapenemase Mechanism",
  subtitle          = "Non-KPC includes OXA-48, NDM, VIM, and other mechanisms",
  legend.labs       = c("KPC", "Non-KPC (OXA-48/NDM/VIM/Other)"),
  palette           = c("#374E55", "#DF8F44"),
  xlim              = c(0, 90),
  break.x.by        = 15,
  ylim              = c(0, 1),
  surv.scale        = "percent",
  ggtheme           = jama_theme,
  tables.theme      = theme_cleantable()
)
```

### Therapy Effect Stratified by Carbapenemase Mechanism {#fig-km4}

```{r km-plot4, fig.height=6.5, fig.cap="Kaplan-Meier survival curves: therapy × carbapenemase interaction"}
surv4 <- surv_infected |>
  filter(carbapenemase != "Not typed") |>
  mutate(
    strata_label = factor(
      paste0(therapy_group, "\n(", carba_group, ")"),
      levels = c(
        "Novel BL/BLI\n(KPC)",
        "Novel BL/BLI\n(Non-KPC (OXA-48/NDM/VIM/Other))",
        "Other therapies\n(KPC)",
        "Other therapies\n(Non-KPC (OXA-48/NDM/VIM/Other))"
      )
    )
  )

fit4 <- survfit(Surv(surv_time, surv_event) ~ strata_label, data = surv4)

ggsurvplot(
  fit4,
  data              = surv4,
  pval              = TRUE,
  pval.size         = 3.5,
  conf.int          = TRUE,
  conf.int.alpha    = 0.10,
  risk.table        = TRUE,
  risk.table.height = 0.28,
  risk.table.title  = "No. at risk",
  risk.table.fontsize = 2.8,
  xlab              = "Days from CRE Infection",
  ylab              = "Overall Survival",
  title             = "D  Therapy Effect Stratified by Carbapenemase Mechanism",
  subtitle          = "Novel BL/BLI vs other therapies within KPC and Non-KPC groups",
  legend.labs       = c("Novel BL/BLI — KPC",
                        "Novel BL/BLI — Non-KPC",
                        "Other therapies — KPC",
                        "Other therapies — Non-KPC"),
  palette           = c("#374E55", "#80A0A8", "#DF8F44", "#F2C27A"),
  xlim              = c(0, 90),
  break.x.by        = 15,
  ylim              = c(0, 1),
  surv.scale        = "percent",
  ggtheme           = jama_theme,
  tables.theme      = theme_cleantable()
)
```

::: callout-note
## Interpretation Notes

- **Novel BL/BLI** = ceftazidime-avibactam, meropenem-vaborbactam, or cefiderocol (any exposure during infection episode). Patients may also have received other agents concurrently.
- **Time origin** = date of CRE infection; censored at hospital discharge or 90 days.
- **Carbapenemase groups:** KPC (n=`r sum(surv_infected$carba_group=="KPC", na.rm=TRUE)`); Non-KPC (n=`r sum(surv_infected$carba_group=="Non-KPC (OXA-48/NDM/VIM/Other)", na.rm=TRUE)`); not typed (n=`r sum(surv_infected$carbapenemase=="Not typed", na.rm=TRUE)`, excluded from plots C and D).
- These are **observational, unadjusted** survival curves. Differences may reflect confounding by infection severity, source, or era.
:::

## Adjusted Cox Proportional Hazards Model

To account for potential confounding by study period, disease severity, age, and infection source, a multivariable Cox model was fitted with therapy group (Novel BL/BLI vs other) as the primary exposure.

```{r cox-model}
# ── Build model dataset ───────────────────────────────────────────────────────
cox_data <- surv_infected |>
  mutate(
    infect_src = factor(
      case_when(
        infection_source___1 == 1                              ~ "Bloodstream",
        infection_source___2 == 1                              ~ "Respiratory",
        infection_source___3 == 1                              ~ "Urinary",
        infection_source___4 == 1 | infection_source___5 == 1 ~ "Surgical/Abd",
        infection_source___6 == 1                              ~ "Other",
        TRUE                                                   ~ NA_character_
      ),
      levels = c("Bloodstream", "Respiratory", "Urinary", "Surgical/Abd", "Other")
    ),
    therapy_fac = relevel(factor(therapy_group), ref = "Other therapies"),
    period_fac  = relevel(factor(period),         ref = "Retrospective"),
    meld_s = as.numeric(scale(meld_score)),   # standardised (per SD)
    age_s  = as.numeric(scale(age))
  ) |>
  filter(!is.na(meld_score), !is.na(infect_src))

# ── Fit model ─────────────────────────────────────────────────────────────────
cox_fit <- coxph(
  Surv(surv_time, surv_event) ~
    therapy_fac + period_fac + meld_s + age_s + infect_src,
  data = cox_data
)

# ── Extract HR table ──────────────────────────────────────────────────────────
cox_tbl <- broom::tidy(cox_fit, exponentiate = TRUE, conf.int = TRUE) |>
  mutate(
    label = recode(term,
      "therapy_facNovel BL/BLI"    = "Novel BL/BLI vs Other",
      "period_facProspective"      = "Prospective vs Retrospective",
      "meld_s"                     = "MELD score (per SD)",
      "age_s"                      = "Age (per SD)",
      "infect_srcRespiratory"      = "Respiratory vs Bloodstream",
      "infect_srcUrinary"          = "Urinary vs Bloodstream",
      "infect_srcSurgical/Abd"     = "Surgical/Abd vs Bloodstream",
      "infect_srcOther"            = "Other source vs Bloodstream"
    ),
    sig = ifelse(p.value < 0.05, "*", "")
  ) |>
  select(label, HR = estimate, `95% CI lower` = conf.low,
         `95% CI upper` = conf.high, `p value` = p.value)

knitr::kable(cox_tbl, digits = c(0, 2, 2, 2, 3),
  caption = paste0("Adjusted Cox proportional hazards model: all-cause mortality ",
                   "from CRE infection (n = ", nrow(cox_data),
                   ", events = ", sum(cox_data$surv_event), ")"))
```

### Forest Plot {#fig-forest}

```{r forest-plot, fig.height=5.5, fig.cap="Forest plot of adjusted hazard ratios from Cox proportional hazards model"}
forest_data <- cox_tbl |>
  rename(HR = HR, lower = `95% CI lower`, upper = `95% CI upper`, p = `p value`) |>
  mutate(
    label = factor(label, levels = rev(label)),
    group = case_when(
      grepl("Novel|Prospective", label) ~ "Primary",
      grepl("MELD|Age", label)          ~ "Continuous",
      TRUE                               ~ "Infection source"
    ),
    sig     = p < 0.05,
    hr_text = sprintf("%.2f (%.2f\u2013%.2f)", HR, lower, upper),
    p_text  = ifelse(p < 0.001, "<0.001",
               ifelse(p < 0.01,  sprintf("%.3f", p),
                                  sprintf("%.2f",  p)))
  )

ggplot(forest_data, aes(x = HR, y = label)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50",
             linewidth = 0.5) +
  geom_errorbar(aes(xmin = lower, xmax = upper, colour = group),
                orientation = "y", width = 0.25, linewidth = 0.7) +
  geom_point(aes(colour = group, shape = sig), size = 3.5) +
  # HR and p-value columns (positioned within scale range)
  geom_text(aes(x = 20, label = hr_text),
            hjust = 1, size = 3.0, family = "mono") +
  geom_text(aes(x = 28, label = p_text),
            hjust = 1, size = 3.0, family = "mono") +
  # Column headers
  annotate("text", x = 20, y = 9.35, label = "HR (95% CI)",
           hjust = 1, size = 3.2, fontface = "bold") +
  annotate("text", x = 28, y = 9.35, label = "p",
           hjust = 1, size = 3.2, fontface = "bold") +
  # Group dividers
  annotate("segment", x = 0.12, xend = 5.5, y = 5.5, yend = 5.5,
           colour = "grey80", linewidth = 0.3) +
  annotate("segment", x = 0.12, xend = 5.5, y = 3.5, yend = 3.5,
           colour = "grey80", linewidth = 0.3) +
  # Vertical divider before text columns
  annotate("segment", x = 6.5, xend = 6.5, y = 0.5, yend = 9,
           colour = "grey70", linewidth = 0.3) +
  # Group labels
  annotate("text", x = 0.13, y = 8.75, label = "Treatment & Period",
           hjust = 0, size = 2.9, fontface = "bold.italic", colour = "grey35") +
  annotate("text", x = 0.13, y = 5.2,  label = "Patient characteristics",
           hjust = 0, size = 2.9, fontface = "bold.italic", colour = "grey35") +
  annotate("text", x = 0.13, y = 3.2,
           label = "Infection source (ref: Bloodstream)",
           hjust = 0, size = 2.9, fontface = "bold.italic", colour = "grey35") +
  scale_x_continuous(
    trans  = "log",
    limits = c(0.12, 30),
    breaks = c(0.25, 0.5, 1, 2, 4),
    labels = c("0.25", "0.5", "1", "2", "4"),
    name   = "Hazard Ratio (log scale)"
  ) +
  scale_colour_manual(
    values = c("Primary"          = "#374E55",
               "Continuous"       = "#DF8F44",
               "Infection source" = "#B24745"),
    guide = "none"
  ) +
  scale_shape_manual(
    values = c("FALSE" = 16, "TRUE" = 18),
    guide  = "none"
  ) +
  labs(
    title    = "Adjusted Cox Proportional Hazards Model",
    subtitle = paste0("Outcome: all-cause mortality from CRE infection",
                      "  |  n = ", nrow(cox_data),
                      ", events = ", sum(cox_data$surv_event)),
    y = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    plot.title         = element_text(face = "bold", size = 12),
    plot.subtitle      = element_text(size = 9.5, colour = "grey40"),
    axis.text.y        = element_text(size = 9.5),
    axis.text.x        = element_text(size = 9),
    axis.title.x       = element_text(size = 9.5),
    panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.3),
    axis.line.y        = element_blank(),
    axis.ticks.y       = element_blank()
  )
```

### Adjusted Survival Curves {#fig-adj-km}

Survival curves derived from the fitted Cox model with all covariates held at their reference or mean values, varying only therapy group. This removes the confounding effects of study period, disease severity, age, and infection source, giving a cleaner estimate of the therapy effect on survival.

```{r adj-km, fig.width=9, fig.height=5.5, fig.cap="Adjusted survival curves from Cox proportional hazards model. Covariates held at: retrospective period, mean MELD score, mean age, bloodstream infection source. Shaded bands = 95% confidence intervals. Percentages at right = adjusted 90-day survival."}
# ── Covariate grid: mean continuous variables, reference-level factors ─────
covariate_means <- cox_data |>
  summarise(
    meld_s     = mean(meld_s,  na.rm = TRUE),
    age_s      = mean(age_s,   na.rm = TRUE),
    period_fac = "Retrospective",
    infect_src = "Bloodstream"
  )

newdata_adj <- bind_rows(
  covariate_means |> mutate(therapy_fac = "Novel BL/BLI"),
  covariate_means |> mutate(therapy_fac = "Other therapies")
) |>
  mutate(
    therapy_fac = factor(therapy_fac,
                         levels = c("Other therapies", "Novel BL/BLI")),
    period_fac  = factor(period_fac,
                         levels = c("Retrospective", "Prospective")),
    infect_src  = factor(infect_src,
                         levels = c("Bloodstream", "Respiratory", "Urinary",
                                    "Surgical/Abd", "Other"))
  )

# ── Adjusted survfit ─────────────────────────────────────────────────────────
adj_fit <- survfit(cox_fit, newdata = newdata_adj)

tidy_adj <- function(sf, labels) {
  map_df(seq_along(labels), function(i) {
    tibble(time  = sf$time,
           surv  = sf$surv[, i],
           lower = sf$lower[, i],
           upper = sf$upper[, i],
           group = labels[i])
  }) |> mutate(group = factor(group, levels = labels))
}

adj_df <- bind_rows(
  tibble(time = 0, surv = 1, lower = 1, upper = 1,
         group = factor(c("Novel BL/BLI", "Other therapies"),
                        levels = c("Novel BL/BLI", "Other therapies"))),
  tidy_adj(adj_fit, c("Novel BL/BLI", "Other therapies"))
)

# ── Risk table (observed counts) ─────────────────────────────────────────────
risk_tbl <- map_df(seq(0, 90, 15), function(t) {
  surv_infected |>
    group_by(therapy_group) |>
    summarise(at_risk = sum(surv_time >= t, na.rm = TRUE), .groups = "drop") |>
    mutate(time = t)
}) |>
  mutate(therapy_group = factor(therapy_group,
                                levels = c("Novel BL/BLI", "Other therapies")))

# ── 90-day survival labels ────────────────────────────────────────────────────
surv_90 <- adj_df |>
  filter(time <= 90) |>
  group_by(group) |>
  slice_tail(n = 1) |>
  mutate(label = paste0(round(surv * 100, 1), "%"))

pal <- c("Novel BL/BLI" = "#374E55", "Other therapies" = "#DF8F44")

# ── Main curve ────────────────────────────────────────────────────────────────
p_adj_km <- ggplot(adj_df, aes(x = time, y = surv * 100,
                                colour = group, fill = group)) +
  geom_ribbon(aes(ymin = lower * 100, ymax = upper * 100),
              alpha = 0.12, colour = NA) +
  geom_step(linewidth = 0.85) +
  geom_text(data = surv_90,
            aes(x = 92, y = surv * 100, label = label, colour = group),
            hjust = 0, vjust = -0.7, size = 3.2, fontface = "bold",
            show.legend = FALSE) +
  geom_hline(yintercept = 50, linetype = "dotted",
             colour = "grey70", linewidth = 0.4) +
  scale_colour_manual(values = pal, name = NULL) +
  scale_fill_manual(values   = pal, name = NULL) +
  scale_x_continuous(limits  = c(0, 100), breaks = seq(0, 90, 15),
                     expand  = expansion(mult = c(0, 0.01))) +
  scale_y_continuous(limits  = c(0, 105), breaks = seq(0, 100, 25),
                     labels  = function(x) paste0(x, "%"),
                     expand  = expansion(mult = c(0, 0))) +
  labs(
    title    = "Adjusted Survival Curves — Novel BL/BLI vs Other Therapies",
    subtitle = paste0(
      "Cox model adjusted for period, MELD score, age, and infection source\n",
      "Covariates held at: retrospective period, mean MELD & age, bloodstream source"
    ),
    x = "Days from CRE Infection",
    y = "Survival Probability"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(size = 9, colour = "grey40",
                                      lineheight = 1.3),
    legend.position    = c(0.22, 0.18),
    legend.background  = element_rect(fill = alpha("white", 0.8), colour = NA),
    legend.key.width   = unit(1.4, "cm"),
    legend.text        = element_text(size = 10),
    axis.text          = element_text(size = 10),
    axis.title         = element_text(size = 10),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3),
    plot.margin        = margin(8, 60, 0, 8)
  ) +
  coord_cartesian(clip = "off")

# ── Risk table ────────────────────────────────────────────────────────────────
p_risk <- ggplot(risk_tbl,
                 aes(x = time, y = fct_rev(therapy_group), label = at_risk)) +
  geom_text(aes(colour = therapy_group), size = 3.2) +
  scale_colour_manual(values = pal, guide = "none") +
  scale_x_continuous(limits  = c(0, 100), breaks = seq(0, 90, 15)) +
  scale_y_discrete(labels = c("Other therapies" = "Other",
                               "Novel BL/BLI"    = "Novel BL/BLI")) +
  labs(x = NULL, y = NULL, title = "No. at risk") +
  theme_classic(base_size = 10) +
  theme(
    plot.title   = element_text(size = 9, face = "bold", colour = "grey30",
                                margin = margin(b = 2)),
    axis.text.x  = element_blank(),
    axis.ticks   = element_blank(),
    axis.line    = element_blank(),
    axis.text.y  = element_text(size = 9),
    panel.grid   = element_blank(),
    plot.margin  = margin(0, 60, 4, 8)
  )

p_adj_km / p_risk + plot_layout(heights = c(5, 1))
```

### Proportional Hazards Assumption {#sec-ph}

::: {.callout-note collapse="true"}
## Schoenfeld Residuals Test — click to expand

The proportional hazards (PH) assumption was tested for each covariate using scaled Schoenfeld residuals (`cox.zph()`). A non-significant p-value and approximately flat smooth curve indicate the assumption is satisfied.

```{r ph-test}
ph_test <- cox.zph(cox_fit)

# Formal test table
ph_tbl <- as.data.frame(ph_test$table) |>
  tibble::rownames_to_column("Covariate") |>
  mutate(
    Covariate = recode(Covariate,
      "therapy_fac"  = "Novel BL/BLI vs Other",
      "period_fac"   = "Prospective vs Retrospective",
      "meld_s"       = "MELD score (per SD)",
      "age_s"        = "Age (per SD)",
      "infect_src"   = "Infection source",
      "GLOBAL"       = "GLOBAL"
    ),
    Verdict = case_when(
      Covariate == "GLOBAL" ~ ifelse(p < 0.05, "⚠ Violated",  "✅ Satisfied"),
      TRUE                  ~ ifelse(p < 0.05, "⚠ Violated",  "✅ Satisfied")
    )
  ) |>
  rename(`χ²` = chisq, df = df, `p value` = p)

knitr::kable(ph_tbl, digits = c(0, 3, 0, 3, 0),
  caption = "Schoenfeld residuals test for proportional hazards assumption") |>
  kableExtra::row_spec(nrow(ph_tbl), bold = TRUE)
```

```{r ph-plot, fig.height=7, fig.cap="Scaled Schoenfeld residuals by covariate. Approximately flat smooth curves indicate the PH assumption is satisfied."}
ph_gg <- ggcoxzph(ph_test,
  var     = c("therapy_fac", "period_fac", "meld_s", "age_s", "infect_src"),
  ggtheme = theme_classic(base_size = 10) +
    theme(
      plot.title         = element_text(face = "bold", size = 10),
      axis.text          = element_text(size = 8, colour = "grey30"),
      axis.title         = element_text(size = 9),
      panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3)
    )
)

clean_labels <- c(
  "Novel BL/BLI vs Other  (p = 0.241)",
  "Prospective vs Retro  (p = 0.491)",
  "MELD score  (p = 0.196)",
  "Age  (p = 0.949)",
  "Infection source  (p = 0.079)"
)

for (i in seq_along(ph_gg)) {
  ph_gg[[i]] <- ph_gg[[i]] +
    labs(title = clean_labels[i], x = "Time (days)", y = NULL) +
    geom_hline(yintercept = 0, linetype = "dashed",
               colour = "grey50", linewidth = 0.4)
}

patchwork::wrap_plots(ph_gg, ncol = 2) +
  patchwork::plot_annotation(
    title    = "Proportional Hazards Assumption — Scaled Schoenfeld Residuals",
    subtitle = paste0(
      "Global test: \u03c7\u00b2 = ",
      round(ph_test$table["GLOBAL", "chisq"], 2),
      "  df = ", ph_test$table["GLOBAL", "df"],
      "  p = ", round(ph_test$table["GLOBAL", "p"], 3),
      "  |  Smooth line should be approximately horizontal if PH holds"
    ),
    theme = theme(
      plot.title    = element_text(face = "bold", size = 12),
      plot.subtitle = element_text(size = 9, colour = "grey40")
    )
  ) &
  theme(axis.title.y = element_blank())
```

**Interpretation:** The PH assumption is satisfied for all covariates individually and globally (global p = 0.205). The infection source term is the weakest (p = 0.079) but remains non-significant. No time-varying coefficients or stratification are required.
:::

::: callout-warning
## Residual Confounding

The prospective cohort period and novel BL/BLI use are correlated by design (novel agents became available over time). Even after adjustment, the HR for novel BL/BLI (HR `r sprintf("%.2f", cox_tbl$HR[cox_tbl$label == "Novel BL/BLI vs Other"])`, 95% CI `r sprintf("%.2f–%.2f", cox_tbl[["95% CI lower"]][cox_tbl$label == "Novel BL/BLI vs Other"], cox_tbl[["95% CI upper"]][cox_tbl$label == "Novel BL/BLI vs Other"])`) and the HR for prospective period both remain significant. This collinearity should be interpreted with caution; the two effects may not be fully separable in this dataset.
:::

## Technical Notes

**Data Source:** CRECOOLT study database (N=692 liver transplant patients)

**Statistical Methods:**

- Chi-square test (or Fisher's exact test where cell counts < 5) for categorical variables
- Wilcoxon rank-sum test for continuous screening counts
- Antibiotic use proportions are exposure-level (patients may appear multiple times)
- Combination therapy proportions are patient-level (first record per patient)

**Software:** R version 4.5.1, with tidyverse and patchwork packages

**Report Generated:** `r format(Sys.Date(), "%B %d, %Y")`
