---
title: "Novel vs. Older Regimens: Recurrence & Resistance"
subtitle: "Within-era, IPTW-weighted competing-risks analysis; whole-cohort estimates are confounded by calendar era and differential ascertainment"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(haven)
library(survival)
library(gt)

pal <- c("Old" = "#EE6677", "Novel" = "#228833")

# Formatter for a hazard/subdistribution ratio from a (cox) model
fmt_cox <- function(m, term = "regimenNovel") {
  s <- summary(m); ci <- s$conf.int[term, c("exp(coef)", "lower .95", "upper .95")]
  p <- s$coefficients[term, ncol(s$coefficients)]
  sprintf("%.2f (%.2f–%.2f), p%s", ci[1], ci[2], ci[3],
          ifelse(p < 0.001, "<0.001", paste0("=", formatC(p, format = "f", digits = 3))))
}
```

## Key finding

Among liver transplant recipients with CRE infection, crude rates of **recurrence** and
**emergent resistance** look much higher with novel agents than with older regimens. That
contrast does **not** survive proper analysis. Two structural problems make the whole-cohort
comparison misleading — novel agents were unavailable before 2018 (a positivity violation), and
these outcomes were incompletely recorded in the retrospective era. Restricting to the
contemporaneous **2018–2025** era, where both regimens were used and documented identically, and
applying inverse-probability-of-treatment weighting (IPTW) with death as a competing risk,
**novel and older regimens show no detectable difference** in either endpoint.

::: {.callout-note}
## How to read this page

The valid comparison — the **2018–2025 era IPTW analysis** — is presented first. The
whole-cohort competing-risks models appear afterward as a **secondary, uncorrected** analysis,
because they are confounded by calendar era and by differential outcome ascertainment (documented
below). Death is treated as a competing event throughout.
:::

## Background

Two distinct threats to validity apply to this question:

1. **Survival bias / competing risk of death.** Patients on older regimens die earlier and more
   often, so they have less opportunity to recur or develop resistance. Treating death as ordinary
   censoring biases the comparison; the appropriate tools are **cause-specific Cox** and
   **Fine–Gray subdistribution** models.
2. **Calendar era — confounding and ascertainment.** Novel agents entered use only from 2018, so a
   whole-cohort treatment contrast conflates *drug* with *time period*. As shown below, the two
   eras also recorded these outcomes differently.

::: {.callout-note}
## Why death is treated as a competing risk

Death is **not** the outcome here — recurrence and emergent resistance are. But both can only
occur in a living patient: once someone dies, they can never subsequently recur or develop
resistance. An event that makes the event of interest impossible is, by definition, a *competing
risk*.

Standard survival methods (Kaplan–Meier, ordinary Cox) treat death as ordinary **censoring**,
which assumes the patient is still at risk and could experience the event later. For a patient who
has died that assumption is false — their future probability of recurrence is zero, not unknown —
so treating deaths as censored **overestimates** the incidence of recurrence and resistance. This
matters acutely here because mortality is much higher in the older-regimen arm (≈48% vs ≈27% in
the full cohort): censoring those deaths would inflate that arm's apparent event risk the most.
Modeling death as a competing event (Fine–Gray for cumulative incidence; cause-specific Cox for
the rate) is precisely the correction for the **survival bias** that motivated this analysis.

*Note: the competing-risks correction alone does not overturn the crude imbalance — the era
restriction and the ascertainment differences below are what explain it.*
:::

```{r load-data}
raw <- read_sav("CRECOOLT_onlyinfections.sav")

cr <- raw |>
  transmute(
    record_id,
    regimen = factor(if_else(as.numeric(new_drug) == 1, "Novel", "Old"), levels = c("Old", "Novel")),
    study_period = as.numeric(study_period),
    inf_date    = as.Date(cre_infection_date),
    death       = as.numeric(death),
    death_date  = as.Date(death_date),
    dischg_date = as.Date(discharge_date),
    clear_date  = as.Date(clearence_date),
    end_date    = as.Date(end_date),
    rec   = as.numeric(cre_recurrence),
    res   = as.numeric(resistance),
    t_rec = as.numeric(CREinf_to_relapse),
    t_res = as.numeric(CREinf_to_resistance),
    age   = as.numeric(age),
    meld  = as.numeric(meld_score),
    multisite = as.numeric(multisite_colonization),
    sofa  = as.numeric(sofa)
  ) |>
  mutate(
    t_death  = as.numeric(death_date - inf_date),
    # discard discharge dates recorded before infection (data-entry errors),
    # fall back to the latest other valid contact date for censoring
    fu_dischg = if_else(as.numeric(dischg_date - inf_date) >= 0, as.numeric(dischg_date - inf_date), NA_real_),
    fu_clear  = if_else(as.numeric(clear_date  - inf_date) >= 0, as.numeric(clear_date  - inf_date), NA_real_),
    fu_end    = if_else(as.numeric(end_date    - inf_date) >= 0, as.numeric(end_date    - inf_date), NA_real_),
    last_valid = pmax(fu_dischg, fu_clear, fu_end, na.rm = TRUE),
    last_valid = if_else(is.finite(last_valid), last_valid, NA_real_)
  )

# Build (time, status) for a competing-risks endpoint:
#   status 1 = event of interest, 2 = death without event, 0 = censored
make_cr <- function(d, evflag, evtime) {
  d |>
    mutate(
      status = case_when(.data[[evflag]] == 1 ~ 1L, death == 1 ~ 2L, TRUE ~ 0L),
      time = case_when(.data[[evflag]] == 1 ~ .data[[evtime]], death == 1 ~ t_death, TRUE ~ last_valid),
      time = if_else(time == 0, 0.5, time)
    ) |>
    filter(!is.na(time), is.finite(time), time > 0)
}

d_rec <- make_cr(cr, "rec", "t_rec")
d_res <- make_cr(cr, "res", "t_res")
```

The cohort comprises `r nrow(raw)` liver transplant recipients with CRE infection
(`r sum(cr$regimen == "Old")` older regimen, `r sum(cr$regimen == "Novel")` novel). Event times
come from the dataset's `CREinf_to_relapse` and `CREinf_to_resistance` fields; death timing from
`death_date`. Discharge dates recorded before the infection date
(`r sum(as.numeric(cr$dischg_date - cr$inf_date) < 0, na.rm = TRUE)` records) were treated as
invalid, with censoring falling back to the latest valid contact date.

## Why a whole-cohort comparison would mislead

### Positivity: novel agents only exist from 2018

```{r positivity}
table(`Study period` = factor(cr$study_period, 1:2, c("2010–2017", "2018–2025")),
      Regimen = cr$regimen) |>
  as.data.frame.matrix() |>
  rownames_to_column("Study period") |>
  gt() |>
  tab_header(title = "Treatment by study period",
             subtitle = "Novel regimens appear only from 2018 — a structural non-overlap")
```

No patient before 2018 could receive a novel agent, so a whole-cohort contrast is partly a
comparison of eras rather than of treatments. A causal comparison must restrict to the period
where both regimens were genuinely used.

### Differential outcome ascertainment {#sec-ascertainment}

Before interpreting any cross-era difference, we ask whether recurrence and resistance were
recorded *the same way* in both periods. They were not.

```{r}
#| label: tbl-ascertainment
#| tbl-cap: "Outcome and follow-up documentation by study era. Recurrence and resistance flags carry no missing values in either era — retrospective patients are uniformly coded \"No\" — yet the supporting follow-up dates are almost entirely absent before 2018."

era_lab <- function(x) factor(as.numeric(x), 1:2, c("2010–2017", "2018–2025"))

raw |>
  mutate(era = era_lab(study_period),
         relapse_date = as.Date(relapse_date),
         new_resistance_date = as.Date(new_resistance_date),
         clear_date = as.Date(clearence_date)) |>
  group_by(era) |>
  summarise(
    N = n(),
    `Recurrence = Yes`      = sprintf("%d (%.1f%%)", sum(cre_recurrence == 1, na.rm = TRUE), 100 * mean(cre_recurrence == 1, na.rm = TRUE)),
    `Resistance = Yes`      = sprintf("%d (%.1f%%)", sum(resistance == 1, na.rm = TRUE), 100 * mean(resistance == 1, na.rm = TRUE)),
    `Flag missing (NA)`     = sum(is.na(cre_recurrence)) + sum(is.na(resistance)),
    `Clearance date`        = sprintf("%d (%.0f%%)", sum(!is.na(clear_date)), 100 * mean(!is.na(clear_date))),
    `Resistance date`       = sum(!is.na(new_resistance_date)),
    .groups = "drop"
  ) |>
  gt(rowname_col = "era") |>
  tab_source_note(md("*Counts of \"Yes\" outcomes and presence of follow-up dates. Percentages are within-era.*"))
```

Three features mark this as differential ascertainment rather than a true era effect:

- **Implausible event rates.** Recorded recurrence is `r sprintf("%.1f%%", 100 * mean(raw$cre_recurrence[as.numeric(raw$study_period) == 1] == 1, na.rm = TRUE))`
  retrospectively vs `r sprintf("%.1f%%", 100 * mean(raw$cre_recurrence[as.numeric(raw$study_period) == 2] == 1, na.rm = TRUE))` prospectively, and **zero** resistance
  events were recorded among `r sum(as.numeric(raw$study_period) == 1)` retrospective patients.
- **Uniform "No" coding, not missingness.** Neither flag has any `NA`; every retrospective
  patient carries an explicit "No", the signature of a chart abstraction that defaulted absent
  information to negative.
- **Follow-up documentation collapses.** Not one retrospective patient has a recorded
  microbiological clearance date — yet recurrence and resistance can only be detected through
  repeat cultures and follow-up susceptibility testing.

::: {.callout-important}
## Consequence for interpretation

The near-zero retrospective rates almost certainly reflect **incomplete outcome ascertainment**,
not a genuine era difference. Older-regimen patients appear event-free partly because their
events were not captured. Consequently the **whole-cohort crude rates and competing-risks models
(shown later) are confounded by ascertainment** and should be read as uncorrected. The
**2018–2025 within-era IPTW analysis** below is the only comparison in which both arms were
documented identically.
:::

## Primary analysis — 2018–2025 era with IPTW

### Propensity model and covariate balance

```{r iptw}
era2 <- cr |> filter(study_period == 2, !is.na(age)) |>
  mutate(novel = as.integer(regimen == "Novel"))

# Parsimonious propensity model (baseline severity)
ps_fit <- glm(novel ~ age + meld + multisite + sofa, family = binomial, data = era2)
era2$ps <- predict(ps_fit, type = "response")
pT <- mean(era2$novel)
era2$iptw <- ifelse(era2$novel == 1, pT / era2$ps, (1 - pT) / (1 - era2$ps))

# Standardized mean differences before/after weighting
smd <- function(x, g, w = rep(1, length(x))) {
  m1 <- weighted.mean(x[g == 1], w[g == 1]); m0 <- weighted.mean(x[g == 0], w[g == 0])
  v1 <- weighted.mean((x[g == 1] - m1)^2, w[g == 1]); v0 <- weighted.mean((x[g == 0] - m0)^2, w[g == 0])
  (m1 - m0) / sqrt((v1 + v0) / 2)
}
covs <- c("age", "meld", "multisite", "sofa")
tibble(Covariate = covs,
       `SMD (unweighted)` = map_dbl(covs, ~smd(era2[[.x]], era2$novel)),
       `SMD (IPTW)`       = map_dbl(covs, ~smd(era2[[.x]], era2$novel, era2$iptw))) |>
  mutate(across(where(is.numeric), ~round(., 3))) |>
  gt() |>
  tab_header(title = "Covariate balance",
             subtitle = "Stabilized IPTW; |SMD| < 0.1 indicates good balance")
```

The N in this era is `r nrow(era2)` (`r sum(era2$regimen == "Old")` older,
`r sum(era2$regimen == "Novel")` novel). Stabilized weights are well-behaved (range
`r sprintf("%.2f–%.2f", min(era2$iptw), max(era2$iptw))`) and propensity overlap is complete, so
positivity holds within the era.

### Weighted competing-risks estimates

```{r era-models}
era_fg <- function(d, label, weighted) {
  d$ev <- factor(d$status, 0:2, c("censor", label, "death"))
  fg <- finegray(Surv(time, ev) ~ ., data = d[, c("time", "ev", "regimen", "iptw", "record_id")],
                 etype = label)
  w <- if (weighted) fg$fgwt * fg$iptw else fg$fgwt
  coxph(Surv(fgstart, fgstop, fgstatus) ~ regimen + cluster(record_id),
        weights = w, data = fg, robust = TRUE)
}
e_rec   <- make_cr(era2, "rec", "t_rec")
e_res   <- make_cr(era2, "res", "t_res")
er_recU <- era_fg(e_rec, "recurrence", FALSE)
er_recW <- era_fg(e_rec, "recurrence", TRUE)
er_resU <- era_fg(e_res, "resistance", FALSE)
er_resW <- era_fg(e_res, "resistance", TRUE)
```

Within the contemporaneous era, the IPTW-weighted Fine–Gray subdistribution hazard ratio
(novel vs. older) is **`r fmt_cox(er_recW)`** for CRE recurrence and **`r fmt_cox(er_resW)`** for
resistance emergence — neither significant. Weighting changes the estimates only marginally from
the unweighted era models (recurrence `r fmt_cox(er_recU)`; resistance `r fmt_cox(er_resU)`),
because the arms are already well balanced within the era. The decisive correction is the era
restriction, not the weighting.

### Weighted cumulative incidence {#sec-cif}

```{r}
#| label: fig-cif
#| fig-cap: "IPTW-weighted cumulative incidence of CRE recurrence and resistance emergence within the 2018–2025 era, with death as a competing risk. Shaded bands are 95% confidence intervals; markers show the 90-day estimate with its 95% CI."
#| fig-width: 9
#| fig-height: 4.5

grid <- 0:120
wcif_band <- function(d, label, nice) {
  d$ev <- factor(d$status, 0:2, c("censor", label, "death"))
  sf <- survfit(Surv(time, ev) ~ regimen, data = d, weights = iptw)
  k <- which(sf$states == label)
  s <- summary(sf, times = grid, extend = TRUE)
  tibble(time = s$time, est = s$pstate[, k], lower = s$lower[, k], upper = s$upper[, k],
         regimen = sub("regimen=", "", as.character(s$strata)), endpoint = nice)
}
band <- bind_rows(
  wcif_band(e_rec, "recurrence", "CRE recurrence"),
  wcif_band(e_res, "resistance", "Resistance emergence")
) |> mutate(regimen = factor(regimen, levels = c("Old", "Novel")))
pt90 <- band |> filter(time == 90)

ggplot(band, aes(time, est, color = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 90, linetype = "dashed", colour = "grey60", linewidth = 0.3) +
  geom_errorbar(data = pt90, aes(ymin = lower, ymax = upper), width = 4,
                position = position_nudge(x = rep(c(-1.5, 1.5), length.out = nrow(pt90)))) +
  geom_point(data = pt90, size = 2.2,
             position = position_nudge(x = rep(c(-1.5, 1.5), length.out = nrow(pt90)))) +
  facet_wrap(~ endpoint) +
  scale_color_manual(values = pal) + scale_fill_manual(values = pal) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = c(0, 30, 60, 90, 120)) +
  labs(x = "Days from CRE infection", y = "Cumulative incidence",
       color = "Regimen", fill = "Regimen") +
  theme_minimal(base_size = 12) + theme(legend.position = "top")
```

The weighted curves overlap for both endpoints, consistent with the null subdistribution hazard
ratios. Late-time steps rest on few patients still at risk and should not be over-interpreted.

## Secondary analysis — full cohort (uncorrected)

::: {.callout-warning}
## These estimates are confounded

The models in this section pool 2010–2025. They are reported for transparency only. Because novel
agents did not exist before 2018 and recurrence/resistance were incompletely ascertained in the
retrospective era, the large apparent excess with novel regimens is substantially an artifact of
era and ascertainment, **not** evidence of harm.
:::

```{r crude}
bind_rows(
  d_rec |> count(regimen, status) |> mutate(endpoint = "CRE recurrence"),
  d_res |> count(regimen, status) |> mutate(endpoint = "Resistance emergence")
) |>
  mutate(status = factor(status, 0:2, c("Censored", "Event", "Death (competing)"))) |>
  pivot_wider(names_from = status, values_from = n, values_fill = 0) |>
  select(endpoint, regimen, Event, `Death (competing)`, Censored) |>
  gt(groupname_col = "endpoint", rowname_col = "regimen") |>
  tab_header(title = "Crude competing-risks event counts by regimen (full cohort)")
```

```{r full-models}
# Cause-specific Cox: death treated as censoring
cs_rec <- coxph(Surv(time, status == 1) ~ regimen, data = d_rec)
cs_res <- coxph(Surv(time, status == 1) ~ regimen, data = d_res)

# Fine-Gray (unadjusted and multisite-adjusted) via finegray + weighted Cox
fg_fit <- function(d, label, adjust = FALSE) {
  d$ev <- factor(d$status, 0:2, c("censor", label, "death"))
  d$multisite <- factor(if_else(d$multisite == 1, "Yes", "No"), levels = c("No", "Yes"))
  rhs <- if (adjust) c("regimen", "multisite") else "regimen"
  fg <- finegray(reformulate(".", response = "Surv(time, ev)"),
                 data = d[, c("time", "ev", rhs)], etype = label)
  coxph(reformulate(rhs, response = "Surv(fgstart, fgstop, fgstatus)"),
        weights = fgwt, data = fg)
}
fg_rec  <- fg_fit(d_rec, "recurrence")
fg_res  <- fg_fit(d_res, "resistance")
fg_recA <- fg_fit(d_rec, "recurrence", adjust = TRUE)
fg_resA <- fg_fit(d_res, "resistance", adjust = TRUE)
```

In the pooled cohort the Fine–Gray SHR is `r fmt_cox(fg_rec)` for recurrence and
`r fmt_cox(fg_res)` for resistance, and adjustment for multisite colonization barely changes them
(`r fmt_cox(fg_recA)` and `r fmt_cox(fg_resA)`). These large "effects" disappear within the
contemporaneous era, identifying them as era/ascertainment artifacts.

## All model variants {#sec-summary}

```{r}
#| label: tbl-models
#| tbl-cap: "Effect of novel vs. older regimens on CRE recurrence and resistance emergence across model specifications. Reference = older regimen. The 2018–2025 era rows are the valid comparison; full-cohort rows are confounded by era and ascertainment."

tribble(
  ~Cohort, ~Model, ~`CRE recurrence`, ~`Resistance emergence`,
  "2018–2025 era (valid)", "Fine–Gray SHR, IPTW-weighted", fmt_cox(er_recW), fmt_cox(er_resW),
  "2018–2025 era (valid)", "Fine–Gray SHR, unadjusted",    fmt_cox(er_recU), fmt_cox(er_resU),
  "Full cohort (confounded)", "Fine–Gray SHR, unadjusted",   fmt_cox(fg_rec),  fmt_cox(fg_res),
  "Full cohort (confounded)", "Fine–Gray SHR, + multisite",  fmt_cox(fg_recA), fmt_cox(fg_resA),
  "Full cohort (confounded)", "Cause-specific Cox HR",       fmt_cox(cs_rec),  fmt_cox(cs_res)
) |>
  gt(groupname_col = "Cohort", rowname_col = "Model") |>
  cols_label(`CRE recurrence` = md("**CRE recurrence**<br>Novel vs Old"),
             `Resistance emergence` = md("**Resistance emergence**<br>Novel vs Old")) |>
  tab_footnote("Stabilized IPTW from a propensity model (age, MELD, multisite colonization, SOFA); robust standard errors.",
               locations = cells_stub(rows = "Fine–Gray SHR, IPTW-weighted")) |>
  tab_footnote("Cause-specific Cox treats death as censoring (effect on the event-specific hazard).",
               locations = cells_stub(rows = "Cause-specific Cox HR")) |>
  tab_source_note(md("*SHR, subdistribution hazard ratio; HR, hazard ratio.*"))
```

## Interpretation

After accounting for death as a competing risk and restricting to the contemporaneous era with
IPTW, novel agents show **no detectable increase** in CRE recurrence or resistance. The crude and
whole-cohort excess is most consistent with cross-era differences — including demonstrable
**differential outcome ascertainment** in 2010–2017 — rather than a causal effect of the newer
drugs.

::: {.callout-warning}
## Caveats

- The within-era older-regimen arm is small (events: `r sum(e_rec$status == 1 & e_rec$regimen == "Old")`
  recurrence, `r sum(e_res$status == 1 & e_res$regimen == "Old")` resistance), so the null findings
  carry wide confidence intervals — evidence against a large harmful effect, not proof of equivalence.
- IPTW addresses *measured* confounding only; residual confounding by indication cannot be excluded.
- The ascertainment asymmetry should be confirmed with the study team (were retrospective
  follow-up cultures and repeat susceptibility testing systematically performed?).
:::

**Bottom line.** Lead with the 2018–2025 IPTW analysis: novel and older regimens are
indistinguishable for recurrence and resistance once death, calendar era, and ascertainment are
handled. The headline crude imbalance is an artifact.

```{r session, include=FALSE}
sessionInfo()
```
